---
title: "Statistical Analysis Plan: Palmer Penguins"
subtitle: "PHB 243b Project 1"
author: "[Your Names Here]"
date: last-modified
# ============================================================
# RENDERING INSTRUCTIONS
# ============================================================
# This document can be rendered as either PDF or slides.
#
# Option 1: Render to PDF (recommended for submission)
#   quarto render analysis_plan_template.qmd --to pdf
#
# Option 2: Render to slides (for presentation)
#   quarto render analysis_plan_template.qmd --to revealjs
#
# Option 3: Set a default format by uncommenting ONE line below:
# format: pdf
# format: revealjs
#
# Option 4: Render both formats at once:
#   quarto render analysis_plan_template.qmd
# ============================================================
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
    geometry:
      - margin=1in
    fontsize: 11pt
  revealjs:
    theme: serif
    slide-number: true
    smaller: true
    scrollable: true
    toc: true
    toc-depth: 1
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(palmerpenguins)
library(knitr)
```

# Introduction

## Background and Rationale

<!--
PROMPT: Provide 2-3 paragraphs describing:
- The scientific context (Antarctic penguin ecology, morphometrics)
- Why this analysis is important or interesting
- Brief mention of relevant prior work (cite Gorman et al. 2014)

Example opening: "Understanding the relationship between morphometric
measurements in penguins provides insight into..."
-->

[Your background text here]

## Research Questions

<!--
PROMPT: State your primary research question clearly. Consider:
- What outcome are you trying to explain or predict?
- What predictors will you examine?
- Are you interested in species differences?

Be specific. A weak question: "What predicts body mass?"
A stronger question: "To what extent does flipper length predict body
mass in Palmer penguins, and does this relationship differ by species?"
-->

**Primary Research Question:**

[Your primary question here]

**Secondary Questions (if any):**

1. [Secondary question 1]
2. [Secondary question 2]

## Hypotheses

<!--
PROMPT: State your hypotheses before looking at the data.
- What relationships do you expect to find?
- What is the biological rationale for these expectations?
- State both null and alternative hypotheses if appropriate.
-->

[Your hypotheses here]

# Data Description

## Data Source

<!--
PROMPT: Describe where the data come from. Include:
- Original data collector (Dr. Kristen Gorman)
- Research station (Palmer Station LTER, Antarctica)
- Time period of data collection
- How you are accessing the data (palmerpenguins R package)
- Citation for the data
-->

[Your data source description here]

**Citation:** Horst AM, Hill AP, Gorman KB (2020). palmerpenguins: Palmer
Archipelago (Antarctica) penguin data. R package version 0.1.0.

## Study Population

<!--
PROMPT: Describe the study population:
- What species are included?
- What islands were sampled?
- What years were data collected?
- Any inclusion/exclusion criteria?
-->

[Your study population description here]

## Sample Size

<!--
PROMPT: Report expected sample size:
- Total observations in the dataset
- Expected complete cases after handling missing data
- Sample size by species (if relevant)
-->

```{r}
#| label: sample-size
penguins |>
  group_by(species) |>
  summarise(
    n_total = n(),
    n_complete = sum(complete.cases(
      bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g, sex
    ))
  ) |>
  kable(caption = "Sample sizes by species")
```

[Your interpretation of sample sizes here]

# Variables

## Primary Outcome

<!--
PROMPT: Define your primary outcome variable:
- Variable name and units
- How it was measured
- Expected range of values
- Any transformations planned
-->

| Attribute | Description |
|-----------|-------------|
| Variable name | [e.g., body_mass_g] |
| Definition | [e.g., Body mass in grams] |
| Type | [Continuous/Categorical] |
| Units | [e.g., grams] |
| Expected range | [e.g., 2700-6300 g] |
| Transformations | [e.g., None / Log] |

## Primary Predictors

<!--
PROMPT: List and define each predictor variable.
For each, specify name, units, measurement method, rationale.
-->

### Predictor 1: [Variable Name]

| Attribute | Description |
|-----------|-------------|
| Variable name | [e.g., flipper_length_mm] |
| Definition | [e.g., Flipper length in mm] |
| Type | [Continuous/Categorical] |
| Rationale | [Why include?] |

### Predictor 2: [Variable Name]

| Attribute | Description |
|-----------|-------------|
| Variable name | |
| Definition | |
| Type | |
| Rationale | |

### Predictor 3: [Variable Name]

| Attribute | Description |
|-----------|-------------|
| Variable name | |
| Definition | |
| Type | |
| Rationale | |

## Effect Modifiers / Covariates

<!--
PROMPT: List variables that may modify the predictor-outcome
relationship or that you will adjust for (species, sex, island, year).
Explain why each is a modifier vs. predictor.
-->

[Your effect modifiers/covariates here]

# Statistical Methods

## Descriptive Statistics

<!--
PROMPT: Specify what descriptive statistics you will compute:
- For continuous variables (mean, SD, median, IQR, range?)
- For categorical variables (counts, percentages?)
- Will you stratify by species or other groupings?
-->

**Continuous variables:** [mean, SD, median, IQR, range]

**Categorical variables:** [frequencies, percentages]

**Stratification:**

- [ ] Overall
- [ ] By species
- [ ] By sex
- [ ] Other: _______________

## Missing Data

<!--
PROMPT: Describe your approach to missing data:
- How many observations have missing values?
- Strategy: complete case analysis, imputation, etc.
- Justify your choice.
-->

```{r}
#| label: missing-data
penguins |>
  summarise(across(everything(), ~sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing") |>
  filter(Missing > 0) |>
  kable(caption = "Variables with missing values")
```

**Strategy:** [Your strategy and justification here]

## Primary Analysis

<!--
PROMPT: Describe your primary analysis approach:
- What statistical method? (e.g., linear regression)
- Model specification?
- Sequential model building?
-->

### Model Specification

**Model 1:** Simple linear regression

$$\text{body\_mass} = \beta_0 + \beta_1 \times \text{flipper\_length} + \epsilon$$

**Model 2:** Multiple regression

$$\text{body\_mass} = \beta_0 + \beta_1 \times \text{flipper} + \beta_2 \times \text{bill\_length} + \epsilon$$

**Model 3:** Species-adjusted

[Your model equation here]

**Model 4:** With interactions

[Your model equation here]

### Model Comparison Criteria

- [ ] R-squared / Adjusted R-squared
- [ ] RMSE
- [ ] AIC / BIC
- [ ] F-tests for nested models

## Significance Level

- **Alpha:** [e.g., 0.05]
- **Multiple comparison adjustment:** [None / Bonferroni / FDR]

## Diagnostic Checks

| Assumption | Diagnostic Method | If Violated |
|------------|-------------------|-------------|
| Linearity | Residual vs. fitted | Transform |
| Normality | Q-Q plot | |
| Homoscedasticity | Scale-location plot | |
| Independence | Study design | |
| Influential points | Cook's distance | |
| Multicollinearity | VIF | |

# Planned Results

## Table 1: Sample Characteristics

| Characteristic | Adelie | Chinstrap | Gentoo | Overall |
|----------------|--------|-----------|--------|---------|
| **n** | | | | |
| **Bill length (mm)** | | | | |
| Mean (SD) | | | | |
| **Bill depth (mm)** | | | | |
| Mean (SD) | | | | |
| **Flipper length (mm)** | | | | |
| Mean (SD) | | | | |
| **Body mass (g)** | | | | |
| Mean (SD) | | | | |
| **Sex, n (%)** | | | | |
| Female | | | | |
| Male | | | | |

## Table 2: Correlation Matrix

| | Bill Length | Bill Depth | Flipper | Body Mass |
|-------------|-------------|------------|---------|-----------|
| Bill Length | 1.00 | | | |
| Bill Depth | | 1.00 | | |
| Flipper | | | 1.00 | |
| Body Mass | | | | 1.00 |

## Table 3: Model Comparison

| Term | Model 1 | Model 2 | Model 3 | Model 4 |
|------|---------|---------|---------|---------|
| Intercept | | | | |
| Flipper length | | | | |
| Bill length | | | | |
| Bill depth | | | | |
| Species | | | | |
| **R-squared** | | | | |
| **RMSE** | | | | |
| **AIC** | | | | |

## Planned Figures

### Figure 1: Species Distribution

- **Type:** Bar chart
- **X-axis:** Species
- **Y-axis:** Count
- **Purpose:** Show sample composition

### Figure 2: Predictor vs. Outcome

- **Type:** Scatter plot
- **X-axis:** Flipper length
- **Y-axis:** Body mass
- **Color:** Species
- **Purpose:** Visualize primary relationship

### Figure 3: Correlation Heatmap

- **Type:** Heatmap
- **Variables:** All numeric
- **Purpose:** Show variable relationships

### Figure 4: Model Diagnostics

- **Type:** 2x2 panel (residuals, Q-Q, scale-location, leverage)
- **Purpose:** Assess model assumptions

# Sensitivity Analyses

<!--
PROMPT: Describe robustness checks:
- Alternative missing data handling
- Excluding outliers
- Alternative model specifications
- Stratified analyses
-->

| Analysis | Rationale | Expected Impact |
|----------|-----------|-----------------|
| Exclude outliers | Test robustness | |
| Multiple imputation | Compare to complete case | |
| Stratify by species | Check pooling validity | |
| Log-transform outcome | Address non-normality | |

# Limitations

## Scope and Generalizability

[Results specific to Palmer Archipelago population]

## Causal Inference

[Observational data, cannot establish causation]

## Data Limitations

[Limited time period 2007-2009, potential measurement error]

## Methodological Limitations

[Model assumptions, variable selection]

# Team Responsibilities

| Task | Responsible | Deadline |
|------|-------------|----------|
| Data cleaning | | |
| Descriptive statistics | | |
| Primary analysis | | |
| Figures | | |
| Write-up | | |
| Review/editing | | |

# References

1. Gorman KB, Williams TD, Fraser WR (2014). Ecological sexual
   dimorphism and environmental variability within a community
   of Antarctic penguins (genus *Pygoscelis*). *PLOS ONE* 9(3): e90081.

2. Horst AM, Hill AP, Gorman KB (2020). palmerpenguins: Palmer
   Archipelago (Antarctica) penguin data. R package.

3. [Additional references]

# Appendix: Session Information

```{r}
#| label: session-info
#| echo: true
sessionInfo()
```

<!--
CHECKLIST before submitting:

[ ] Clear, specific research question(s)
[ ] Complete data description with citations
[ ] All variables defined with units and rationale
[ ] Detailed statistical methods with model specifications
[ ] Missing data strategy with justification
[ ] Table shells showing exactly what will be reported
[ ] Figure descriptions with clear purposes
[ ] Sensitivity analyses planned
[ ] Limitations acknowledged
[ ] All references cited

Goal: Someone else could execute your plan and get the same results.
-->
