---
title: "Reproducible R Development with Docker and renv"
author: "PHB 243"
date: "`r Sys.Date()`"
format: revealjs
---

# Introduction
- Ensuring reproducibility in R Markdown can be challenging.
- Package version mismatches and OS differences cause issues.
- Solution: Use **renv** (package management) + **Docker** (containerization).

---

# Workflow Overview
1. Developer 1: Sets up **renv** & Docker, pushes to GitHub & DockerHub.
2. Developer 2: Clones repo, pulls Docker image, runs & modifies analysis.
3. Uses **bind mounting** to ensure local files persist.

---

# Step 1: Developer 1 - Setup GitHub Repository
```sh
git clone https://github.com/username/penguins-analysis.git
cd penguins-analysis
```

---

# Step 2: Developer 1 - Initialize renv
```r
install.packages("renv")
renv::init()
```

- Creates **renv.lock** to track package versions.
- Ensures package consistency across systems.

---

# Step 3: Developer 1 - Install Required Packages
```r
install.packages("ggplot2")
install.packages("palmerpenguins")
renv::snapshot()
```

- **renv::snapshot()** saves exact package versions.

---

# Step 4: Developer 1 - Create Dockerfile
```dockerfile
FROM rocker/r-ver:4.1.0
WORKDIR /workspace
RUN R -e "install.packages('renv', repos='https://cloud.r-project.org')"
COPY renv.lock renv/activate.R /workspace/
RUN R -e "renv::restore()"
CMD ["/bin/bash"]
```

- **Does NOT include peng1.Rmd** (handled locally by Developer 2).

---

# Step 5: Developer 1 - Build & Push Docker Image
```sh
docker build -t username/penguins-analysis:v1 .
docker login
docker push username/penguins-analysis:v1
```

- Image now available on **DockerHub** for Developer 2.

---

# Step 6: Developer 1 - Push to GitHub
```sh
git add .
git commit -m "Initial renv setup and Docker environment (without Rmd)"
git push origin main
```

- Repository is now accessible to Developer 2.

---

# Step 7: Developer 2 - Clone & Pull Docker Image
```sh
git clone https://github.com/username/penguins-analysis.git
cd penguins-analysis
docker pull username/penguins-analysis:v1
```

---

# Step 8: Developer 2 - Run Docker Interactively
```sh
docker run --rm -it -v "$(pwd):/workspace" -w /workspace username/penguins-analysis:v1 /bin/bash
```

- Mounts **local repository** to `/workspace` inside the container.

---

# Step 9: Developer 2 - Write peng1.Rmd
```rmd
```{r}
library(ggplot2)
library(palmerpenguins)

ggplot(penguins, aes(x = flipper_length_mm, y = bill_length_mm)) +
  geom_point() + theme_minimal()
```
```

---

# Step 10: Developer 2 - Render R Markdown
```r
rmarkdown::render("peng1.Rmd")
```

- Output **peng1.html** is saved **locally** (due to bind mount).

---

# Step 11: Developer 2 - Extend Analysis & Push to GitHub
```rmd
## Body Mass vs. Bill Length
```{r}
ggplot(penguins, aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point() + theme_minimal()
```
```

```sh
git add peng1.Rmd
git commit -m "Added second plot: Body Mass vs. Bill Length"
git push origin main
```

---

# Conclusion
✅ **renv** ensures package reproducibility.  
✅ **Docker** ensures OS & R version consistency.  
✅ **GitHub** & **DockerHub** enable collaboration.  
✅ **Bind mounting** ensures local modifications persist.  

---

# Thank You! 🚀
- Questions?  
- Try setting up your own Dockerized R environment!

