---
title: "Palmer Penguins Analysis - Bill Depth vs Body Mass"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load required R packages
library(palmerpenguins)
library(ggplot2)
library(dplyr)
library(reticulate)

# Clean data for plotting
data(penguins)
penguins_clean <- penguins %>% 
  drop_na(bill_depth_mm, body_mass_g, species)

# Prepare data for Python - convert factors to characters and ensure standard data frame
py_penguins <- penguins_clean %>%
  mutate(species = as.character(species)) %>%
  as.data.frame()
```

## Overview

This document demonstrates how to create the same visualization using both R and Python within a single R Markdown document. We'll visualize the relationship between bill depth and body mass for different penguin species from the Palmer Penguins dataset.

## R Visualization

Using ggplot2 in R to create a scatter plot of bill depth versus body mass, colored by penguin species:

```{r r-plot, fig.width=8, fig.height=5}
# Create scatter plot with R
ggplot(penguins_clean, aes(x = body_mass_g, y = bill_depth_mm, color = species)) +
  geom_point(alpha = 0.8, size = 3) +
  scale_color_manual(values = c("Adelie" = "darkorange", "Chinstrap" = "purple", "Gentoo" = "#008080")) +
  labs(
    title = "Bill Depth vs Body Mass by Penguin Species (R)",
    x = "Body Mass (g)",
    y = "Bill Depth (mm)",
    color = "Species"
  ) +
  theme_minimal()
```

## Python Visualization

Using matplotlib in Python with the pyplot interface:

```{python python-pyplot, fig.width=8, fig.height=5, error=TRUE}
# Import libraries
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt

# Access the penguins data from R
penguins = r.py_penguins

# Define colors for each species
colors = {"Adelie": "darkorange", "Chinstrap": "purple", "Gentoo": "#008080"}

# Create figure with specified size
plt.figure(figsize=(8, 5))

# Create scatter plot for each species using pyplot interface
for species in penguins['species'].unique():
    # Filter data for this species
    species_data = penguins[penguins['species'] == species]
    
    # Plot this species using pyplot directly
    plt.scatter(
        species_data['body_mass_g'],
        species_data['bill_depth_mm'],
        c=colors[species],
        label=species,
        alpha=0.8,
        s=80
    )

# Add title and labels using pyplot methods
plt.title("Bill Depth vs Body Mass by Penguin Species (Python)")
plt.xlabel("Body Mass (g)")
plt.ylabel("Bill Depth (mm)")

# Add legend
plt.legend(title="Species")

# Add grid for better readability
plt.grid(True, linestyle='--', alpha=0.7)

# Adjust layout and show
plt.tight_layout()
plt.show()
```

