---
title: "Multi-Language Data Science"
subtitle: "Combining R and Python in RMarkdown"
author: "phb 243b"
date: "March 12, 2025"
format: 
  revealjs:
    theme: simple
    slide-number: true
    code-fold: false
    incremental: false
    transition: slide
    background-transition: fade
    highlight-style: github
    footer: "Multi-Language Data Science with R and Python"
---

## Why Use Multiple Languages?

- **Leverage strengths** of different ecosystems
- **Reuse existing code** without rewriting
- **Collaborate** across language preferences
- **Choose optimal tools** for each task
- **Learn new languages** while using familiar ones

::: notes
- R excels at statistical analysis and visualization
- Python has strong machine learning and data manipulation libraries
- Some tasks might be easier in one language than the other
:::

## RMarkdown: The Multi-Language Notebook

- Seamlessly combine R, Python, SQL, Julia, and more
- Uses the `reticulate` package for R-Python integration
- Each code chunk can use a different language engine
- Data objects can be shared between environments
- Rendered into HTML, PDF, Word, presentations, etc.

::: notes
- RMarkdown is built on knitr and pandoc
- Quarto is the next generation of RMarkdown
- The reticulate package is what makes the Python integration possible
:::

## Setting Up the Environment {.smaller}

```{r}
#| echo: true
#| eval: false

# In R, load the necessary packages
library(reticulate)  # Core package for R-Python integration
library(ggplot2)     # For R visualizations
library(dplyr)       # For R data manipulation

# Configure Python (optional but recommended)
use_python("/path/to/python")  # Specify Python installation
use_virtualenv("my-env")       # Or use a virtual environment
use_condaenv("my-conda-env")   # Or use a conda environment
```

::: notes
- It's important to be explicit about which Python environment to use
- This helps with reproducibility and avoiding conflicts
- You can specify Python versions and environments in the YAML header too
:::

## Accessing R Objects from Python {.smaller}

```{python}
#| echo: true
#| eval: false

# The 'r' object provides access to the R environment
penguins = r.penguins_clean  # Access an R dataframe

# Check the data
print(type(penguins))
print(penguins.columns)
print(penguins.head())

# Use the data with Python libraries
import matplotlib.pyplot as plt
import numpy as np

# The data is now ready for Python analysis
```

::: notes
- The 'r' object is automatically created by reticulate
- It provides access to any objects in the R global environment
- Type conversions are handled automatically in most cases
:::

## Example: Penguin Visualization in R {.smaller}

```{r}
#| echo: true
#| eval: false

# Create scatter plot with R and ggplot2
ggplot(penguins_clean, aes(x = body_mass_g, y = bill_depth_mm, color = species)) +
  geom_point(alpha = 0.8, size = 3) +
  scale_color_manual(values = c("Adelie" = "darkorange", 
                               "Chinstrap" = "purple", 
                               "Gentoo" = "#008080")) +
  labs(
    title = "Bill Depth vs Body Mass by Penguin Species (R)",
    x = "Body Mass (g)",
    y = "Bill Depth (mm)",
    color = "Species"
  ) +
  theme_minimal()
```

::: notes
- ggplot2 provides a grammar of graphics approach
- We explicitly map colors to species for consistency
- This is the declarative approach common in R
:::

## Example: Penguin Visualization in Python {.smaller}

```{python}
#| echo: true
#| eval: false

# Create scatter plot with Python and matplotlib
import matplotlib.pyplot as plt

# Define colors for each species
colors = {"Adelie": "darkorange", "Chinstrap": "purple", "Gentoo": "#008080"}

# Create figure and axis
fig, ax = plt.subplots(figsize=(8, 5))

# Create scatter plot for each species
for species in penguins['species'].unique():
    species_data = penguins[penguins['species'] == species]
    
    ax.scatter(
        species_data['body_mass_g'],
        species_data['bill_depth_mm'],
        c=colors[species],
        label=species,
        alpha=0.8,
        s=80
    )

ax.set_title("Bill Depth vs Body Mass by Penguin Species (Matplotlib)")
ax.set_xlabel("Body Mass (g)")
ax.set_ylabel("Bill Depth (mm)")
ax.legend(title="Species")
ax.grid(True, linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()
```

::: notes
- Matplotlib takes a more imperative approach
- We need to iterate through species groups manually
- The result is similar but the approach is different
:::

## Passing Objects Between Languages {.smaller}

:::: {.columns}

::: {.column width="50%"}
**R to Python**
```{r}
#| echo: true
#| eval: false
# Create in R
mtcars_subset <- mtcars %>%
  filter(cyl == 8) %>%
  select(mpg, hp, wt)
```

```{python}
#| echo: true
#| eval: false
# Use in Python
cars = r.mtcars_subset
print(cars.mean())
```
:::

::: {.column width="50%"}
**Python to R**
```{python}
#| echo: true
#| eval: false
# Create in Python
import numpy as np
import pandas as pd
py_data = pd.DataFrame({
  'x': np.random.normal(0, 1, 100),
  'y': np.random.normal(5, 2, 100)
})
```

```{r}
#| echo: true
#| eval: false
# Use in R
py_data <- py$py_data
cor(py_data$x, py_data$y)
```
:::

::::

::: notes
- Data types are automatically converted between languages
- R data frames become pandas DataFrames
- Python objects are accessed via the py variable in R
:::

## Data Type Conversions {.smaller}

| R Type | Python Type |
|--------|-------------|
| vector | numpy array |
| data.frame | pandas.DataFrame |
| matrix | numpy.ndarray |
| list | list or dict |
| environment | dict |
| function | function |
| NULL | None |


