---
title: "Statistical Analysis Plan: ADNI MCI Conversion Prediction"
subtitle: "PHB 243b Project 2"
author: "[Your Name Here]"
date: last-modified
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
    geometry:
      - margin=1in
    fontsize: 11pt
  revealjs:
    theme: serif
    slide-number: true
    smaller: true
    scrollable: true
    toc: true
    toc-depth: 1
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(knitr)
# library(ADNIMERGE2)  # Uncomment when data available
```

# Introduction

## Background and Rationale

<!--
PROMPT: Provide 2-3 paragraphs describing:
- Alzheimer's disease as a public health problem
- The role of MCI as a transitional state
- Why predicting MCI-to-AD conversion is clinically important
- Brief mention of prior work (cite Grassi et al. 2019)

Example opening: "Alzheimer's disease (AD) is the most common cause of
dementia, affecting an estimated 6.7 million Americans..."
-->

[Your background text here]

## Research Questions

<!--
PROMPT: State your primary research question clearly. Consider:
- What outcome are you trying to predict? (MCI to AD conversion)
- What predictors will you examine? (baseline clinical/cognitive measures)
- What is the prediction horizon? (3-year follow-up)

Be specific. Example: "Can baseline sociodemographic characteristics,
clinical scales, and neuropsychological test scores predict conversion
from MCI to AD dementia within 3 years?"
-->

**Primary Research Question:**

[Your primary question here]

**Secondary Questions (if any):**

1. Which predictor categories (sociodemographic, clinical, neuropsychological)
   contribute most to prediction accuracy?
2. Does prediction performance differ between EMCI and LMCI subgroups?

## Hypotheses

<!--
PROMPT: State your hypotheses before analyzing the data.
- What variables do you expect to be most predictive?
- What is the clinical/biological rationale?
- What level of discrimination (AUROC) do you expect?

Reference Grassi et al. (2019): AUROC = 0.88
-->

**Primary Hypothesis:**

[Your hypothesis here, e.g., "Baseline ADAS-Cog-13, CDR-SB, and RAVLT
forgetting scores will significantly predict MCI-to-AD conversion..."]

**Expected Performance:**

Based on Grassi et al. (2019), we expect AUROC in the range of 0.75-0.85.

# Data Description

## Data Source

<!--
PROMPT: Describe where the data come from. Include:
- ADNI study description
- ADNIMERGE2 R package
- Data Use Agreement requirements
- Data download date
-->

Data for this analysis come from the Alzheimer's Disease Neuroimaging
Initiative (ADNI), accessed via the ADNIMERGE2 R package (Donohue et al.,
2026). ADNI is a longitudinal multicenter study designed to develop
clinical, imaging, genetic, and biochemical biomarkers for early detection
and tracking of Alzheimer's disease.

**Data access:** [adni.loni.usc.edu](https://adni.loni.usc.edu)

**Data download date:** [Your download date]

**Citation:** Donohue MC, et al. (2026). Alzheimer's clinical research data
via R packages: The alzverse. *Alzheimer's Dement*, 22:e71152.

## Study Population

<!--
PROMPT: Describe the study population:
- Inclusion criteria: MCI diagnosis at baseline (EMCI or LMCI)
- Minimum follow-up: 3 years (or at least one follow-up visit)
- Exclusion criteria: Missing key predictors, no follow-up data
-->

**Inclusion Criteria:**

1. Baseline diagnosis of Mild Cognitive Impairment (EMCI or LMCI)
2. At least 3 years of follow-up OR documented conversion to AD
3. Complete data on primary predictor variables

**Exclusion Criteria:**

1. Baseline diagnosis of CN or Dementia
2. No follow-up visits after baseline
3. Missing data on >50% of predictor variables

## Sample Size

<!--
PROMPT: Report expected sample size:
- Total MCI subjects in ADNIMERGE2
- Expected sample after applying inclusion/exclusion criteria
- Number of converters vs. stable
-->

```{r}
#| label: sample-size
#| eval: false
# Uncomment when ADNIMERGE2 data available
# data(ADSL)
# ADSL |>
#   filter(DX_BL %in% c("EMCI", "LMCI")) |>
#   count(DX_BL) |>
#   kable(caption = "MCI subjects by baseline diagnosis")
```

**Expected Sample Size:**

| Group | N (expected) | Percentage |
|-------|--------------|------------|
| Stable MCI (sMCI) | ~350 | ~64% |
| Converters (cAD) | ~200 | ~36% |
| **Total** | ~550 | 100% |

Based on Grassi et al. (2019) cohort selection criteria.

# Variables

## Primary Outcome

<!--
PROMPT: Define your primary outcome variable:
- Conversion to AD dementia (binary: Yes/No)
- How conversion is defined (any follow-up diagnosis of Dementia)
- Time horizon (within 3 years of baseline)
-->

| Attribute | Description |
|-----------|-------------|
| Variable name | `converter` |
| Definition | Conversion from MCI to AD dementia |
| Type | Binary (0 = Stable, 1 = Converter) |
| Operationalization | Any follow-up diagnosis of "Dementia" within 3 years |
| Reference | Grassi et al. (2019) |

## Primary Predictors

### Sociodemographic Variables

| Variable | Definition | Type | Rationale |
|----------|------------|------|-----------|
| AGE | Age at baseline (years) | Continuous | Age is primary AD risk factor |
| PTEDUCAT | Years of education | Continuous | Cognitive reserve proxy |
| PTGENDER | Sex | Binary | Sex differences in AD risk |
| PTMARRY | Marital status | Categorical | Social support proxy |

### Clinical Scales

| Variable | Definition | Type | Rationale |
|----------|------------|------|-----------|
| CDRSB | CDR Sum of Boxes (0-18) | Continuous | Clinical staging |
| FAQ | Functional Activities Questionnaire (0-30) | Continuous | Functional impairment |

### Neuropsychological Tests

| Variable | Definition | Type | Rationale |
|----------|------------|------|-----------|
| MMSE | Mini-Mental State Exam (0-30) | Continuous | Global cognition |
| ADAS13 | ADAS-Cog 13-item (0-85) | Continuous | AD-specific cognition |
| RAVLT_immediate | RAVLT sum of trials 1-5 | Continuous | Verbal learning |
| RAVLT_learning | RAVLT trial 5 - trial 1 | Continuous | Learning rate |
| RAVLT_forgetting | RAVLT trial 5 - delayed | Continuous | Memory retention |
| LDELTOTAL | Logical Memory delayed recall | Continuous | Episodic memory |
| TRABSCOR | Trail Making Test B (seconds) | Continuous | Executive function |

### MCI Subtype

| Variable | Definition | Type | Rationale |
|----------|------------|------|-----------|
| DX_BL | Baseline MCI subtype | Binary | EMCI vs LMCI |

## Excluded Variables

<!--
PROMPT: Explain why imaging and genetic variables are excluded
(following Grassi et al. approach for clinical translatability)
-->

The following variable categories are intentionally **excluded** to maximize
clinical translatability (Grassi et al., 2019):

- **Imaging:** MRI volumes, PET SUVRs
- **Genetic:** APOE genotype
- **Biomarkers:** CSF Abeta, tau, p-tau

**Rationale:** These measures require specialized equipment and expertise
not available in most community clinical settings.

# Statistical Methods

## Descriptive Statistics

**Continuous variables:** Mean (SD), median (IQR), range

**Categorical variables:** Frequencies, percentages

**Stratification:** By conversion status (sMCI vs cAD)

**Comparison tests:**

- Continuous: t-test or Wilcoxon rank-sum
- Categorical: Chi-square or Fisher's exact

## Missing Data

<!--
PROMPT: Describe your approach to missing data
-->

```{r}
#| label: missing-data
#| eval: false
# Uncomment when data available
# mci_data |>
#   summarise(across(everything(), ~sum(is.na(.))/n()*100)) |>
#   pivot_longer(everything(), names_to = "Variable", values_to = "Pct_Missing") |>
#   filter(Pct_Missing > 0) |>
#   arrange(desc(Pct_Missing)) |>
#   kable(caption = "Percentage missing by variable", digits = 1)
```

**Strategy:** [Complete case analysis / Multiple imputation / Specify]

**Justification:** [Your justification here]

## Machine Learning Methods

### Algorithms

We will implement at least three classification algorithms:

| Algorithm | R Package | Key Hyperparameters |
|-----------|-----------|---------------------|
| Logistic Regression | `glm` | None |
| Random Forest | `randomForest` | ntree, mtry |
| Gradient Boosting | `xgboost` | max_depth, eta, nrounds |
| [Optional: Additional] | | |

### Cross-Validation Strategy

<!--
PROMPT: Describe your validation approach
- Train/test split ratio
- Cross-validation folds
- Site-independent split (if possible)
-->

1. **Train/Test Split:** 70% training, 30% testing
2. **Cross-Validation:** 5-fold CV within training set
3. **Site Independence:** [If feasible, split by ADNI site]

### Hyperparameter Tuning

| Algorithm | Tuning Method | Grid/Range |
|-----------|---------------|------------|
| Random Forest | Grid search | mtry: 2-8, ntree: 100-500 |
| XGBoost | Grid search | max_depth: 3-6, eta: 0.01-0.3 |

## Evaluation Metrics

| Metric | Definition | Target |
|--------|------------|--------|
| AUROC | Area under ROC curve | > 0.75 |
| Sensitivity | TP / (TP + FN) | > 0.70 |
| Specificity | TN / (TN + FP) | > 0.70 |
| PPV | TP / (TP + FP) | Report |
| NPV | TN / (TN + FN) | Report |

**Threshold Selection:** Youden's J statistic (maximize sensitivity + specificity)

## Feature Importance

- **Random Forest:** Mean decrease in Gini / Permutation importance
- **Logistic Regression:** Standardized coefficients
- **XGBoost:** Gain-based importance

# Planned Results

## Table 1: Baseline Characteristics by Conversion Status

| Characteristic | Stable MCI | Converters | p-value |
|----------------|------------|------------|---------|
| **N** | | | |
| **Age, mean (SD)** | | | |
| **Education, mean (SD)** | | | |
| **Female, n (%)** | | | |
| **EMCI, n (%)** | | | |
| **MMSE, mean (SD)** | | | |
| **ADAS-Cog-13, mean (SD)** | | | |
| **CDR-SB, mean (SD)** | | | |
| **FAQ, mean (SD)** | | | |
| **RAVLT immediate, mean (SD)** | | | |
| **RAVLT forgetting, mean (SD)** | | | |
| **LDELTOTAL, mean (SD)** | | | |
| **TMT-B, mean (SD)** | | | |

## Table 2: Model Performance Comparison

| Metric | Logistic Reg | Random Forest | XGBoost |
|--------|--------------|---------------|---------|
| AUROC (95% CI) | | | |
| Sensitivity | | | |
| Specificity | | | |
| PPV | | | |
| NPV | | | |
| Accuracy | | | |

## Table 3: Feature Importance Rankings

| Rank | Logistic Reg | Random Forest | XGBoost |
|------|--------------|---------------|---------|
| 1 | | | |
| 2 | | | |
| 3 | | | |
| 4 | | | |
| 5 | | | |

## Planned Figures

### Figure 1: Cohort Flow Diagram

- **Type:** CONSORT-style flow diagram
- **Purpose:** Show sample selection and exclusions

### Figure 2: ROC Curves

- **Type:** Overlaid ROC curves for all models
- **Purpose:** Compare discrimination across algorithms

### Figure 3: Feature Importance Plot

- **Type:** Bar chart or dot plot
- **Purpose:** Visualize top predictors

### Figure 4: Calibration Plot

- **Type:** Predicted vs. observed probabilities
- **Purpose:** Assess calibration of best model

# Sensitivity Analyses

| Analysis | Rationale | Expected Impact |
|----------|-----------|-----------------|
| EMCI only | Test in early-stage MCI | Lower conversion rate |
| LMCI only | Test in late-stage MCI | Higher conversion rate |
| Exclude outliers | Robustness check | Minimal change expected |
| Alternative imputation | Compare to complete case | Similar results expected |
| Different CV folds | Check stability | Similar AUROC expected |

# Limitations

## Scope and Generalizability

- Results specific to ADNI population (highly educated, mostly White)
- May not generalize to community clinic populations
- ADNI uses strict diagnostic criteria

## Methodological Limitations

- Retrospective cohort design
- Cannot establish causation
- 3-year prediction horizon may miss later converters
- No external validation cohort

## Data Limitations

- Missing data on some predictor variables
- Potential selection bias in ADNI enrollment
- Measurement variability across sites

# Timeline

| Week | Task | Deliverable |
|------|------|-------------|
| 1 (Feb 17-21) | Data preparation, cohort selection | Clean dataset |
| 2 (Feb 24-28) | Feature engineering, initial models | Preliminary results |
| 3 (Mar 2-6) | Model tuning, evaluation, report | Final submission |

**Due Date:** March 3, 2026

# References

1. Donohue MC, et al. (2026). Alzheimer's clinical research data via R
   packages: The alzverse. *Alzheimer's Dement*, 22:e71152.

2. Grassi M, et al. (2019). A Clinically-Translatable Machine Learning
   Algorithm for the Prediction of Alzheimer's Disease Conversion.
   *Front Neurol*, 10:756.

3. Petersen RC. (2004). Mild cognitive impairment as a diagnostic entity.
   *J Intern Med*, 256(3):183-194.

4. Rye I, et al. (2022). Predicting conversion to Alzheimer's disease in
   individuals with Mild Cognitive Impairment. *Sci Rep*, 12:15566.

# Appendix: Code Framework

```{r}
#| label: code-framework
#| echo: true
#| eval: false

# ============================================================
# Data Preparation
# ============================================================
library(ADNIMERGE2)
library(tidyverse)
library(caret)

# Load data
data(ADSL)
data(ADQS)

# Select MCI subjects at baseline
mci_baseline <- ADSL |>
  filter(DX_BL %in% c("EMCI", "LMCI"))

# Define conversion status
# [Your code here]

# ============================================================
# Feature Engineering
# ============================================================
# Select predictor variables
predictors <- c("AGE", "PTEDUCAT", "PTGENDER", "CDRSB", "FAQ",
                "MMSE", "ADAS13", "RAVLT_immediate", "RAVLT_forgetting",
                "LDELTOTAL", "TRABSCOR")

# ============================================================
# Model Training
# ============================================================
# Train/test split
set.seed(243)
train_idx <- createDataPartition(mci_data$converter, p = 0.7, list = FALSE)

# Cross-validation setup
ctrl <- trainControl(method = "cv", number = 5,
                     classProbs = TRUE, summaryFunction = twoClassSummary)

# ============================================================
# Model Evaluation
# ============================================================
# ROC curves
library(pROC)
# [Your code here]
```

<!--
CHECKLIST before submitting:

[ ] Clear, specific research question about MCI conversion prediction
[ ] Complete data description with ADNI/ADNIMERGE2 citations
[ ] All predictor variables defined with clinical rationale
[ ] At least 3 ML algorithms specified with hyperparameters
[ ] Cross-validation strategy clearly described
[ ] Evaluation metrics defined with targets
[ ] Table shells for baseline characteristics and model comparison
[ ] Figure descriptions for ROC curves and feature importance
[ ] Sensitivity analyses planned
[ ] Limitations acknowledged
[ ] All references cited

Goal: Someone else could execute your plan and get the same results.
-->
