

---
title: "Machine Learning predictions in ADNI"
author: "R.G. Thomas"
date: "`r Sys.Date()`"
fontsize: 11pt
output:
  pdf_document:
#   toc: true
    number_sections: true
    highlight: tango
    keep_tex: true 
    includes:
      in_header: "/Users/zenn/prj/preamble.tex"
header-includes:
   - \rhead{Machine learning ((~/sandbox/adni-machine-learning)}
bibliography: /Users/zenn/share/library.bib
latex_engine: xelatex
---

```{r  include=FALSE}
options(knitr.kable.NA = '',dplyr.summarise.inform = FALSE, dplyr.print_max = 1e9)
library(pacman)
p_load(kableExtra, janitor, rmarkdown, tidyverse, knitr)

opts_chunk$set(warning = F, message = F, echo = F, fig.width = 3.2,
       fig.height = 3,  results = "asis", dev = "pdf",
       opts_chunk$set(cache.path = "cache/"))
source("~/share/zz.tools.R")
options(scipen = 1, digits = 2)
```

\hspace{4in}
\fbox{
\parbox{3in}{\normalsize 
Notes: 
\begin{itemize}
\item mimic three ADNI ML papers
\begin{itemize}
\item   grassi, 2021-05-20: 
\item read in ADNIMERGE data and started subset of
converters with at least 3 years of followup. 
\item	gill 
\item	dimitriadis 
\end{itemize}
\item idea for first analysis: 
\item random forest analysis of
cognitive data to find enrichment criteria
\item set of gitlab repo: rgthomas@ucsd.edu/www4????
\end{itemize}
}
}
\listoftodos

# Introduction 
[start with a description of AD and the early stages.
Why are we looking for prediction models? consider NIA intro page] A
diagnosis of Alzheimer's Disease usually only occurs following a long
period, as much as 15 years, of disease progression. 

The length of this prodromal phase differs across individuals.
Identifying those at greatest risk for rapid progression is vital,
both for a clinical and a research perspective.i.e. enrichment. 

Consider the ADNI data archive as a resource for assessing/modeling
progression from MCI -- the earliest stage of observable disease
phenotype -- to AD. The ADNI project was designed by Drs. Leon Thal
and Michael Weiner. Initiated in 2005, the project has 

\begin{wrapfigure}{r}[.4\textwidth]{.4\textwidth}%
{Need an account 
on Loni. login: rgthomas@ucsd.edu. password: see "> pass find loni"   } 
\end{wrapfigure}

First key paper is  @Grassi2019. First step is to mimic
table 2 in the paper. 

\includegraphics[width=7in]{/Users/zenn/sandbox/adni-machine-learning/table2.png} 


other references:

Sample reference section reference. @Dimitriadis2020

Sample reference section reference. @Gill2020a

Sample reference section reference. @Haulcy2021

Sample reference section reference. @Katabathula2021

\begin{marginfigure}
\includegraphics[width=3in]{/Users/zenn/prj/images/sudoku-3d-cube.pdf} \caption{test}\label{fig:unnamed-chunk-8}
\end{marginfigure}

# Methods
```sh

Notes on getting data from ADNI
new password .. see pass find loni

goal 1: mimic Grassi table 2. 
1. determine sample size for adnimerge data set. 
step 1.1: read in ADNIMERGE from ./data directory
> df1 = read.csv("./data/ADNIMERGE.csv")
> dim(df1) # should be greater than 550
>  # looking for 550 subjects with MCI and dx followup for at
>  # least 3 years. 

2. find age mean/sd by coverters

step 2: explore DX variable (see ./data/ADNIMERGE_DICT.csv
and ./data/ADNIMERGE_Methods_21030429.pdf

```









# Results



# Code 
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```

# References

```{r cache=F}
# df  %>%  tabyl(dx_bl)
df  %>%  tabyl(dx)
df = read.csv("./data/ADNIMERGE.csv", na.strings="-4")  %>% 
  sel(RID, VISCODE, ORIGPROT, AGE, MMSE, DX, DX_bl, ADAS11, ADAS13, ADASQ4)  %>% 
  select_all(tolower) %>% 
  fil(dx_bl == "LMCI") 
df_cnt = df  %>%  group_by(rid) %>%  summ(N = n()) %>% fil(N >= 3)  # filter for those with 3 or more visits
df2 = df %>%  fil(rid %in% df_cnt$rid)%>%  arrange(rid, viscode) %>%  group_by(rid)
df3 = df2  %>%  summ(aa = any(dx=="Dementia"))
# df2  %>%  tabyl(dx)
# df  %>%  tabyl(origprot)
# df  %>%  tabyl(dx_bl)
# df3  %>%  tabyl(aa)
# foo  %>%  summ(aa = any(dx=="MCI"))
foo = hd(df2, 100)
df3foo = foo  %>%  summ(aa = any(dx=="Dementia"))





# tab1adni = df  %>%  
# tabyl(race) %>%  
#  adorn_totals(where = "row") 
# #adorn_totals(where = "col")    %>% 
# # sel(-percent) 
# tab1adni$race = sub( "\\/","-", tab1adni$race)
# out = kbl(tab1adni,  booktabs = T,escape=FALSE, 
#     caption = "Race categorization for ADNI.") %>%
# 		kable_paper(latex_options = "striped")    
# # zz.figured(tab1adni, width="2.5")
# zz.figured(tab1adni)
# 
# tab1badni = df  %>%  
# tabyl(race, ORIGPROT) %>%  
#  adorn_totals(where = "row") %>%  
# adorn_totals(where = "col")   
# tab1badni$race = sub( "\\/","-", tab1badni$race)
# # sel(-percent) 
# out = kbl(tab1badni,  booktabs = T,escape=FALSE, 
#     caption = "Race categorization for ADNI by protocol.") %>%
# 		kable_paper(latex_options = "striped")    
# # zz.figured(tab1badni, width="2.5in")
# zz.figured(tab1badni)
# 
dict =  read.csv("./data/ADNIMERGE_DICT.csv", na.strings="-4")  %>% 
	sel(FLDNAME, TBLNAME, TEXT, CODE)  


```


