---
title: "Machine Learning predictions in ADNI"
author: "R.G. Thomas"
date: "`r Sys.Date()`"
fontsize: 11pt
output:
  pdf_document:
    number_sections: true
    highlight: tango
    keep_tex: false
    latex_engine: xelatex
---

```{r setup, include=FALSE}
options(
  knitr.kable.NA = "",
  dplyr.summarise.inform = FALSE,
  dplyr.print_max = 1e9
)
library(pacman)
p_load(
  kableExtra, janitor, rmarkdown, tidyverse, knitr,
  caret, pROC, glmnet, randomForest, e1071, gbm, class
)

opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE,
  fig.width = 5, fig.height = 4,
  results = "asis", dev = "pdf"
)
options(scipen = 1, digits = 2)
set.seed(42)
```

## Notes

**Goal**: Replicate the analysis of Grassi et al. (2019), *Frontiers in
Neurology* 10:756, which developed an ensemble-based machine learning
algorithm to predict 3-year conversion from MCI to AD using
sociodemographic, clinical, and neuropsychological predictors from the
ADNIMERGE dataset.

# Introduction

Alzheimer's Disease (AD) is a neurodegenerative disease characterized by
progressive memory loss, cognitive impairment, and general disability. The
progression of AD comprises a long, unnoticed preclinical stage, followed by
a prodromal stage of Mild Cognitive Impairment (MCI) that leads to severe
dementia and eventually death. Only 20--40\% of MCI individuals will progress
to dementia within 3 years following initial diagnosis.

The prompt identification of subjects at high risk for conversion to AD is of
great importance for clinical trial enrichment and early intervention. Grassi
et al. (2019) proposed a clinically translatable machine learning algorithm,
based on non-invasive and easily collectible predictors, to identify MCI
subjects at risk for conversion to AD. The algorithm achieved an AUROC of
0.88 with 77.7\% sensitivity and 79.9\% specificity.

The present analysis replicates the Grassi et al. Table 2 descriptive
statistics and follows their analytical pipeline using available R
implementations of the machine learning techniques they employed.

# Methods

## Data Loading

```{r load-data}
df_raw <- read.csv("./data/ADNIMERGE.csv", na.strings = "-4")

df <- df_raw |>
  dplyr::select(
    RID, VISCODE, SITE, ORIGPROT,
    AGE, PTGENDER, PTEDUCAT, PTMARRY,
    DX, DX_bl,
    CDRSB, ADAS11, ADAS13, ADASQ4, MMSE,
    RAVLT_immediate, RAVLT_learning,
    RAVLT_forgetting, RAVLT_perc_forgetting,
    LDELTOTAL, DIGITSCOR, TRABSCOR, FAQ,
    Years_bl
  ) |>
  rename_with(tolower) |>
  dplyr::filter(dx_bl %in% c("EMCI", "LMCI"))
```

## Inclusion Criteria

Following Grassi et al. (2019), we select MCI subjects (both Early and Late
MCI) with diagnostic follow-up for at least 3 years from baseline. Converters
are defined as subjects whose diagnosis changed to Dementia at any follow-up
assessment within 3 years.

```{r apply-criteria}
followup_duration <- df |>
  group_by(rid) |>
  dplyr::summarize(
    max_years = max(years_bl, na.rm = TRUE),
    .groups = "drop"
  )

rids_3yr <- followup_duration |>
  dplyr::filter(max_years >= 3) |>
  pull(rid)

df_3yr <- df |>
  dplyr::filter(rid %in% rids_3yr)

converter_status <- df_3yr |>
  dplyr::filter(years_bl <= 3) |>
  group_by(rid) |>
  dplyr::summarize(
    converter = any(dx == "Dementia", na.rm = TRUE),
    .groups = "drop"
  )
```

```{r baseline-data}
df_bl <- df_3yr |>
  dplyr::filter(viscode == "bl") |>
  left_join(converter_status, by = "rid") |>
  dplyr::filter(
    !is.na(converter),
    age >= 55, age <= 90,
    mmse >= 24, mmse <= 30
  ) |>
  mutate(
    group = if_else(converter, "Converter", "Non-converter"),
    mci_subtype = if_else(
      dx_bl == "EMCI", "Early", "Late"
    ),
    sex = ptgender
  )
```

```{r sample-counts}
n_total <- nrow(df_bl)
n_conv <- sum(df_bl$converter)
n_nonconv <- n_total - n_conv
pct_conv <- round(100 * n_conv / n_total, 1)
```

The analytic sample includes `r n_total` MCI subjects with at least 3 years
of diagnostic follow-up: `r n_conv` converters (`r pct_conv`\%) and
`r n_nonconv` non-converters. Grassi et al. reported 550 subjects (197
converters, 35.8\%; 353 non-converters, 64.2\%).

## Descriptive Statistics (Table 2 Replication)

```{r table2-continuous, results='asis'}
cont_vars <- c(
  "age", "pteducat", "cdrsb", "adas11", "adas13", "adasq4",
  "mmse", "ravlt_immediate", "ravlt_learning",
  "ravlt_forgetting", "ravlt_perc_forgetting",
  "ldeltotal", "digitscor", "trabscor", "faq"
)
cont_labels <- c(
  "Age", "Years of education", "CDRSB",
  "ADAS11", "ADAS13", "ADASQ4",
  "MMSE", "RAVLT-I", "RAVLT-L",
  "RAVLT-F", "RAVLT-PF",
  "LDT", "DIGIT", "TMTBT", "FAQ"
)

tbl_cont <- map2_dfr(cont_vars, cont_labels, function(v, lab) {
  stats <- df_bl |>
    group_by(group) |>
    dplyr::summarize(
      mean_val = mean(.data[[v]], na.rm = TRUE),
      sd_val = sd(.data[[v]], na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(stat = sprintf("%.2f (%.2f)", mean_val, sd_val)) |>
    dplyr::select(group, stat) |>
    pivot_wider(names_from = group, values_from = stat)

  n_miss <- sum(is.na(df_bl[[v]]))
  pct_miss <- if (n_miss > 0) {
    sprintf("%.2f", 100 * n_miss / n_total)
  } else {
    "/"
  }
  n_miss_str <- if (n_miss > 0) as.character(n_miss) else "/"

  stats |>
    mutate(
      Variable = lab,
      `Missing N` = n_miss_str,
      `Missing pct` = pct_miss,
      .before = 1
    )
})

kbl(tbl_cont, booktabs = TRUE, linesep = "", escape = TRUE,
    col.names = c(
      "Continuous predictors", "N", "\\%",
      "Mean (S.D.)", "Mean (S.D.)"
    ),
    caption = "Descriptive statistics of continuous predictors
    (cf. Grassi et al., 2019, Table 2).") |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 9
  ) |>
  add_header_above(c(
    " " = 1, "Missing" = 2,
    "Non-converters" = 1, "Converters" = 1
  ))
```

```{r table2-categorical, results='asis'}
cat_summary <- function(data, var, levels_vec) {
  map_dfr(levels_vec, function(lev) {
    data |>
      group_by(group) |>
      dplyr::summarize(
        n_lev = sum(.data[[var]] == lev, na.rm = TRUE),
        pct_lev = 100 * mean(.data[[var]] == lev, na.rm = TRUE),
        .groups = "drop"
      ) |>
      mutate(stat = sprintf("%.0f (%.2f)", n_lev, pct_lev)) |>
      dplyr::select(group, stat) |>
      pivot_wider(names_from = group, values_from = stat) |>
      mutate(Category = lev, .before = 1)
  })
}

tbl_sex <- cat_summary(df_bl, "sex", c("Male", "Female")) |>
  mutate(Variable = c("Sex", ""), .before = 1)

tbl_mci <- cat_summary(
  df_bl, "mci_subtype", c("Early", "Late")
) |>
  mutate(Variable = c("Subtype of MCI", ""), .before = 1)

df_marry <- df_bl |>
  dplyr::filter(ptmarry %in% c(
    "Never married", "Married", "Divorced", "Widowed"
  ))

tbl_marry <- cat_summary(
  df_marry, "ptmarry",
  c("Never married", "Married", "Divorced", "Widowed")
) |>
  mutate(
    Variable = c("Marital status", "", "", ""),
    .before = 1
  )

tbl_cat_all <- bind_rows(tbl_sex, tbl_mci, tbl_marry) |>
  dplyr::select(Variable, Category, `Non-converter`, Converter)

kbl(tbl_cat_all, booktabs = TRUE, linesep = "", escape = TRUE,
    col.names = c(
      "Categorical predictors", "",
      "N (\\%)", "N (\\%)"
    ),
    caption = "Descriptive statistics of categorical predictors
    (cf. Grassi et al., 2019, Table 2).") |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 9
  ) |>
  add_header_above(c(
    " " = 2, "Non-converters" = 1, "Converters" = 1
  ))
```

## Feature Transformation and Selection

Following Grassi et al. (2019), continuous variables are standardized
(mean = 0, SD = 1) and categorical variables are dichotomized using
one-hot encoding. Variables correlated above $r = 0.75$ are replaced with
their first principal component.

```{r feature-prep, results='hide'}
pred_cont <- c(
  "age", "pteducat", "cdrsb", "adas11", "adas13", "adasq4",
  "mmse", "ravlt_immediate", "ravlt_learning",
  "ravlt_forgetting", "ravlt_perc_forgetting",
  "ldeltotal", "trabscor", "faq"
)

df_model <- df_bl |>
  dplyr::select(rid, converter, all_of(pred_cont),
                sex, mci_subtype, ptmarry) |>
  dplyr::filter(
    ptmarry %in% c(
      "Never married", "Married", "Divorced", "Widowed"
    )
  ) |>
  mutate(converter = factor(
    converter, levels = c(FALSE, TRUE),
    labels = c("NC", "cAD")
  ))

impute_median <- function(x) {
  x[is.na(x)] <- median(x, na.rm = TRUE)
  x
}

df_model <- df_model |>
  mutate(across(all_of(pred_cont), impute_median))

adas_cor <- cor(
  df_model[, c("adas11", "adas13", "adasq4")],
  use = "complete.obs"
)
adas_pca <- prcomp(
  df_model[, c("adas11", "adas13", "adasq4")],
  scale. = TRUE
)
df_model$adas_pc1 <- adas_pca$x[, 1]

ravltf_cor <- cor(
  df_model[, c("ravlt_forgetting", "ravlt_perc_forgetting")],
  use = "complete.obs"
)
ravltf_pca <- prcomp(
  df_model[, c("ravlt_forgetting", "ravlt_perc_forgetting")],
  scale. = TRUE
)
df_model$ravltf_pc1 <- ravltf_pca$x[, 1]

pred_final <- c(
  "age", "pteducat", "cdrsb", "adas_pc1",
  "mmse", "ravlt_immediate", "ravlt_learning",
  "ravltf_pc1", "ldeltotal", "trabscor", "faq",
  "sex", "mci_subtype", "ptmarry"
)

df_analytic <- df_model |>
  dplyr::select(rid, converter, all_of(pred_final))

dummies <- dummyVars(
  converter ~ sex + mci_subtype + ptmarry,
  data = df_analytic
)
cat_encoded <- predict(dummies, df_analytic) |>
  as.data.frame()
names(cat_encoded) <- make.names(names(cat_encoded))

pred_numeric <- c(
  "age", "pteducat", "cdrsb", "adas_pc1",
  "mmse", "ravlt_immediate", "ravlt_learning",
  "ravltf_pc1", "ldeltotal", "trabscor", "faq"
)

df_final <- bind_cols(
  df_analytic[, c("rid", "converter")],
  df_analytic[, pred_numeric],
  cat_encoded
)

preproc <- preProcess(
  df_final[, pred_numeric],
  method = c("center", "scale")
)
df_final[, pred_numeric] <- predict(
  preproc, df_final[, pred_numeric]
)

feature_names <- setdiff(names(df_final), c("rid", "converter"))
```

The ADAS scores (ADAS11, ADAS13, ADASQ4) exhibited pairwise correlations
above 0.75 and were replaced by their first principal component (ADAS-PC1,
explaining `r round(100 * summary(adas_pca)$importance[2, 1], 1)`\% of
variance). Similarly, RAVLT-F and RAVLT-PF were replaced by RAVLT-F-PC1
(`r round(100 * summary(ravltf_pca)$importance[2, 1], 1)`\% of variance).

## Machine Learning Analysis

We train several supervised ML algorithms following the Grassi methodology:
logistic regression (LR), elastic net (EN), support vector machines with
linear and radial kernels (SVM-Linear, SVM-RBF), random forest (RF),
gradient tree boosting (GBM), and k-nearest neighbors (kNN). We use
repeated 10-fold stratified cross-validation and evaluate via AUROC.

```{r ml-setup, results='hide'}
train_ctrl <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

x_train <- df_final[, feature_names]
y_train <- df_final$converter
```

```{r ml-models, results='hide', cache=TRUE}
fit_lr <- train(
  x = x_train, y = y_train,
  method = "glm", family = "binomial",
  metric = "ROC", trControl = train_ctrl
)

fit_en <- train(
  x = x_train, y = y_train,
  method = "glmnet",
  metric = "ROC", trControl = train_ctrl,
  tuneLength = 10
)

fit_rf <- train(
  x = x_train, y = y_train,
  method = "rf",
  metric = "ROC", trControl = train_ctrl,
  tuneLength = 5
)

fit_svml <- train(
  x = x_train, y = y_train,
  method = "svmLinear",
  metric = "ROC", trControl = train_ctrl
)

fit_svmr <- train(
  x = x_train, y = y_train,
  method = "svmRadial",
  metric = "ROC", trControl = train_ctrl,
  tuneLength = 5
)

fit_gbm <- train(
  x = x_train, y = y_train,
  method = "gbm", verbose = FALSE,
  metric = "ROC", trControl = train_ctrl,
  tuneLength = 5
)

fit_knn <- train(
  x = x_train, y = y_train,
  method = "knn",
  metric = "ROC", trControl = train_ctrl,
  tuneLength = 10
)

models <- list(
  LR = fit_lr, EN = fit_en, RF = fit_rf,
  `SVM-Linear` = fit_svml, `SVM-RBF` = fit_svmr,
  GBM = fit_gbm, kNN = fit_knn
)
```

## Weighted Rank Ensemble

Following Grassi et al., we combine predictions from all trained algorithms
using a weighted average of rescaled ranks, where weights are proportional
to each algorithm's cross-validated AUROC.

```{r ensemble, results='hide'}
cv_aurocs <- map_dbl(models, function(m) {
  max(m$results$ROC)
})

ensemble_probs <- map(models, function(m) {
  predict(m, x_train, type = "prob")[["cAD"]]
})

rank_preds <- map(ensemble_probs, function(p) {
  rank(p) / length(p)
})
rank_mat <- do.call(cbind, rank_preds)

weights <- cv_aurocs / sum(cv_aurocs)
ensemble_score <- as.numeric(rank_mat %*% weights)

roc_ensemble <- roc(
  y_train, ensemble_score,
  levels = c("NC", "cAD"), direction = "<", quiet = TRUE
)
auc_ensemble <- auc(roc_ensemble)
ci_ensemble <- ci.auc(roc_ensemble, method = "bootstrap",
                       boot.n = 2000, quiet = TRUE)
```

# Results

## Model Performance

```{r performance-table, results='asis'}
perf_tbl <- tibble(
  Algorithm = names(cv_aurocs),
  `CV AUROC` = sprintf("%.3f", cv_aurocs)
)

ensemble_row <- tibble(
  Algorithm = "Ensemble",
  `CV AUROC` = sprintf(
    "%.3f (%.3f--%.3f)",
    auc_ensemble, ci_ensemble[1], ci_ensemble[3]
  )
)

perf_tbl <- bind_rows(perf_tbl, ensemble_row)

kbl(perf_tbl, booktabs = TRUE, linesep = "", escape = TRUE,
    caption = "Cross-validated AUROC for each algorithm and the
    weighted rank ensemble. Grassi et al. reported an ensemble
    test AUROC of 0.88 (95\\% CI 0.85--0.91).") |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 10
  )
```

## ROC Curve

```{r roc-plot, results='markup', fig.width=4.5, fig.height=4.5}
roc_list <- map(models, function(m) {
  probs <- predict(m, x_train, type = "prob")[["cAD"]]
  roc(y_train, probs,
      levels = c("NC", "cAD"), direction = "<", quiet = TRUE)
})
roc_list[["Ensemble"]] <- roc_ensemble

cols <- c(
  LR = "blue", EN = "darkgreen", RF = "red",
  `SVM-Linear` = "purple", `SVM-RBF` = "orange",
  GBM = "brown", kNN = "gray50", Ensemble = "black"
)

plot(roc_list[["Ensemble"]], col = "black", lwd = 2,
     main = "ROC Curves", legacy.axes = TRUE)
iwalk(roc_list[names(roc_list) != "Ensemble"], function(r, nm) {
  lines(r, col = cols[nm], lty = 2)
})
legend("bottomright",
       legend = paste0(
         names(roc_list), " (",
         sprintf("%.3f", map_dbl(roc_list, auc)), ")"
       ),
       col = cols[names(roc_list)], lwd = c(rep(1, 7), 2),
       lty = c(rep(2, 7), 1), cex = 0.6, bty = "n")
```

## Sensitivity and Specificity at Various Thresholds

```{r threshold-table, results='asis'}
sens_targets <- c(1.0, 0.975, 0.95, 0.90, 0.85, 0.80, 0.75)
threshold_tbl <- map_dfr(sens_targets, function(s) {
  coord <- coords(roc_ensemble, x = s, input = "sensitivity",
                  ret = c("threshold", "sensitivity",
                          "specificity"),
                  best.method = "closest.topleft")
  sens_actual <- coord$sensitivity
  spec_actual <- coord$specificity
  pred_class <- if_else(
    as.numeric(ensemble_score) >= coord$threshold,
    "cAD", "NC"
  )
  tp <- sum(pred_class == "cAD" & y_train == "cAD")
  fp <- sum(pred_class == "cAD" & y_train == "NC")
  fn <- sum(pred_class == "NC" & y_train == "cAD")
  tn <- sum(pred_class == "NC" & y_train == "NC")
  ppv <- tp / (tp + fp)
  npv <- tn / (tn + fn)
  bal_acc <- (sens_actual + spec_actual) / 2
  f1 <- 2 * (ppv * sens_actual) / (ppv + sens_actual)

  tibble(
    `Target Sens.` = sprintf("%.1f", s),
    `Sens. (actual)` = sprintf("%.1f", 100 * sens_actual),
    `Spec. (pct)` = sprintf("%.1f", 100 * spec_actual),
    `PPV (pct)` = sprintf("%.1f", 100 * ppv),
    `NPV (pct)` = sprintf("%.1f", 100 * npv),
    `Bal. Acc.` = sprintf("%.3f", bal_acc),
    `F1` = sprintf("%.3f", f1)
  )
})

kbl(threshold_tbl, booktabs = TRUE, linesep = "",
    escape = TRUE,
    caption = "Performance at various sensitivity thresholds
    (cf. Grassi et al., 2019, Table 5).") |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 9
  )
```

## Feature Importance

Following Grassi et al., we assess feature importance by fitting
univariate logistic regression models for each predictor and computing
the test AUROC.

```{r feature-importance, results='asis'}
feat_auroc <- map_dfr(feature_names, function(f) {
  roc_obj <- tryCatch({
    mod <- glm(
      converter ~ ., data = data.frame(
        converter = y_train, x = x_train[[f]]
      ),
      family = "binomial"
    )
    pred_prob <- predict(mod, type = "response")
    r <- roc(y_train, pred_prob,
             levels = c("NC", "cAD"), direction = "<",
             quiet = TRUE)
    ci_r <- ci.auc(r, method = "bootstrap",
                    boot.n = 2000, quiet = TRUE)
    tibble(
      Feature = f,
      AUROC = as.numeric(auc(r)),
      CI_lo = as.numeric(ci_r[1]),
      CI_hi = as.numeric(ci_r[3])
    )
  }, error = function(e) {
    tibble(Feature = f, AUROC = NA_real_,
           CI_lo = NA_real_, CI_hi = NA_real_)
  })
  roc_obj
})

feat_auroc <- feat_auroc |>
  dplyr::filter(!is.na(AUROC)) |>
  arrange(desc(AUROC))

feat_tbl <- feat_auroc |>
  mutate(
    AUROC_str = sprintf("%.3f", AUROC),
    CI = sprintf("%.3f--%.3f", CI_lo, CI_hi)
  ) |>
  dplyr::select(Feature, AUROC = AUROC_str, `95\\% CI` = CI)

kbl(feat_tbl, booktabs = TRUE, linesep = "", escape = TRUE,
    caption = "Individual test AUROC of each feature via
    univariate logistic regression (cf. Grassi et al., 2019,
    Table 6).") |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 9
  )
```

```{r feature-importance-plot, results='markup', fig.width=5, fig.height=5}
feat_plot <- feat_auroc |>
  mutate(Feature = fct_reorder(Feature, AUROC))

ggplot(feat_plot, aes(x = AUROC, y = Feature)) +
  geom_point() +
  geom_errorbarh(
    aes(xmin = CI_lo, xmax = CI_hi), height = 0.3
  ) +
  geom_vline(xintercept = 0.5, linetype = "dashed",
             color = "gray50") +
  labs(
    x = "AUROC",
    y = NULL,
    title = "Individual Predictor AUROC",
    subtitle = "with 95% bootstrap CI"
  ) +
  theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank())
```

# Code
```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```

# References

- Grassi, M., Rouleaux, N., Caldirola, D., Loewenstein, D.,
  Schruers, K., Perna, G., \& Dumontier, M. (2019). A novel
  ensemble-based machine learning algorithm to predict the conversion
  from mild cognitive impairment to Alzheimer's disease using
  socio-demographic characteristics, clinical information, and
  neuropsychological measures. *Frontiers in Neurology*, 10, 756.
- Dimitriadis, S.I., et al. (2018). Random forest feature selection,
  fusion and ensemble strategy. *Neurobiology of Aging*, 69, 216--228.
- Gill, S., et al. (2020). Using machine learning to predict dementia
  from neuropsychiatric symptom and neuroimaging data. *Journal of
  Alzheimer's Disease*, 75(1), 277--288.
- Haulcy, R., \& Glass, J. (2021). Classifying Alzheimer's disease
  using audio and text-based representations of speech. *Frontiers in
  Psychology*, 11, 624137.
- Katabathula, S., et al. (2021). Predict Alzheimer's disease using
  hippocampus MRI data. *BMC Medical Informatics and Decision Making*,
  21(Suppl 2), 197.
