---
title: "PHB 243b - Practicum in Biostatistics"
author:
  - name: "Ronald G. Thomas, PhD"
    affiliation: "Instructor"
    email: "rgthomas@ucsd.edu"
  - name: "Man Luo"
    affiliation: "TA"
    email: "maluo@ucsd.edu"
date: "1/6/26"
format:
  revealjs:
    theme: default
    slide-number: true
    transition: slide
    width: 1600
    height: 900
    fontsize: 24pt
---

## Overview

- Survey of **343 collaborative biostatisticians**
- Assessed importance of **16 team science skills**
- Compared across career stages, degrees, and job sectors
- Identified **3 additional skills** from open responses

::: {.notes}
This presentation summarizes key findings from Slade et al. (2023) on essential team science skills for collaborative biostatisticians.
:::

---

## Table 1: Team Science Skills (Part 1)

| Skill | Description |
|-------|-------------|
| **Databases, data sources, and data collection tools** | Understand nuances of working with databases, data sources, and data collection tools, including their advantages/limitations for answering clinical/scientific questions |
| **Developing clinical/scientific domain knowledge** | Learn about a biomedical domain as needed for a given project to critically assess and interpret scientific results |
| **Regulatory requirements** | Understand, implement, and explain relevant data security and regulatory requirements |
| **Institutional structure** | Navigate one's institutional infrastructure, processes, and funding mechanisms |

---

## Table 1: Team Science Skills (Part 2)

| Skill | Description |
|-------|-------------|
| **Statistical analysis plans** | Write and critique statistical analysis plans with sufficient detail to ensure replicable fulfillment of study objectives |
| **Reproducibility** | Utilize a documentation process for project reproducibility including data storage, coding, analysis plans, reports, etc. so that the project can be reproduced by others |
| **Coding** | Write code to manage data and/or implement statistical methods using accurate and efficient coding practices |
| **Literature review** | Conduct review of literature and background information to identify gaps in scientific knowledge to motivate a given project |

---

## Table 1: Team Science Skills (Part 3)

| Skill | Description |
|-------|-------------|
| **Learning new statistical methods** | Learn and implement unfamiliar statistical methods needed for a given project |
| **Professional correspondence** | Correspond with collaborators promptly, appropriately, and clearly |
| **Time/project management** | Complete projects while planning for and managing multiple projects simultaneously |
| **Effective meeting strategy** | Facilitate effective discussions with clinical/scientific collaborators to achieve meeting goals and develop follow-up steps |

---

## Table 1: Team Science Skills (Part 4)

| Skill | Description |
|-------|-------------|
| **Scientific communication** | Effectively communicate scientific/statistical concepts with collaborators |
| **Presenting results** | Present results formally, both verbally and in writing, in a clear and non-technical manner |
| **Using strong statistical voice** | Advocate and negotiate for good and ethical statistical practices including integrating and resolving differing scientific approaches |
| **Collaboration with analytic colleagues** | Navigate a network of colleagues to identify additional methodological expertise and solicit guidance or collaboration as needed |

---

## Table 1: Additional Skills Identified

| Skill | Description |
|-------|-------------|
| **Diversity, equity, inclusion, and accessibility** | Incorporate diverse perspectives and ensure fair representation both in the research process and in the workplace |
| **Professional development** | Actively shape one's career trajectory, including continuing education, setting boundaries, and avoiding burnout |
| **Mentoring and supervision** | Serve as a formal mentor to trainees, a supervisor to employees, or a peer mentor to colleagues with diverse training backgrounds |

---

## Table 3: Importance of Skills (n=343)

```{r}
#| echo: false
#| message: false

library(knitr)
library(kableExtra)

skills_data <- data.frame(
  Skill = c(
    "Databases, data sources, and data collection tools",
    "Developing clinical/scientific domain knowledge",
    "Regulatory requirements",
    "Institutional structure",
    "Statistical analysis plans",
    "Reproducibility",
    "Coding",
    "Literature review",
    "Learning new statistical methods",
    "Professional correspondence",
    "Time/project management",
    "Effective meeting strategy",
    "Scientific communication",
    "Presenting results",
    "Using strong statistical voice",
    "Collaboration with analytic colleagues"
  ),
  `Not Important` = c(
    "8 (2.4%)", "4 (1.2%)", "26 (7.8%)", "32 (9.7%)",
    "4 (1.2%)", "2 (0.6%)", "2 (0.6%)", "35 (10.6%)",
    "3 (0.9%)", "2 (0.6%)", "1 (0.3%)", "2 (0.6%)",
    "0 (0.0%)", "0 (0.0%)", "0 (0.0%)", "3 (0.9%)"
  ),
  `Somewhat Important` = c(
    "37 (11.2%)", "55 (16.6%)", "102 (30.7%)", "126 (38.1%)",
    "16 (4.8%)", "20 (6.0%)", "11 (3.3%)", "102 (30.8%)",
    "37 (11.2%)", "11 (3.3%)", "14 (4.2%)", "40 (12.1%)",
    "4 (1.2%)", "4 (1.2%)", "26 (7.9%)", "45 (13.6%)"
  ),
  Important = c(
    "103 (31.1%)", "149 (44.9%)", "125 (37.7%)", "129 (39.0%)",
    "72 (21.7%)", "114 (34.3%)", "61 (18.4%)", "133 (40.2%)",
    "128 (38.7%)", "98 (29.5%)", "78 (23.6%)", "139 (42.0%)",
    "77 (23.3%)", "66 (20.2%)", "112 (33.8%)", "156 (47.0%)"
  ),
  `Absolutely Essential` = c(
    "183 (55.3%)", "124 (37.3%)", "79 (23.8%)", "44 (13.3%)",
    "240 (72.3%)", "196 (59.0%)", "257 (77.6%)", "61 (18.4%)",
    "163 (49.2%)", "221 (66.6%)", "238 (71.9%)", "150 (45.3%)",
    "250 (75.5%)", "257 (78.6%)", "193 (58.3%)", "128 (38.6%)"
  )
)

kable(skills_data[1:8,], format = "html", col.names = c("Skill", "Not Important", "Somewhat Important", "Important", "Absolutely Essential")) |>
  kable_styling(font_size = 18, bootstrap_options = c("striped", "condensed"))
```

---

## Table 3: Importance of Skills (continued)

```{r}
#| echo: false
#| message: false

kable(skills_data[9:16,], format = "html", col.names = c("Skill", "Not Important", "Somewhat Important", "Important", "Absolutely Essential")) |>
  kable_styling(font_size = 18, bootstrap_options = c("striped", "condensed"))
```

---

## Figure 1: Skills by Career Stage

```{r}
#| echo: false
#| message: false
#| fig-width: 14
#| fig-height: 7

library(ggplot2)
library(tidyr)
library(dplyr)

fig1_data <- data.frame(
  Skill = c(
    "Coding", "Presenting results", "Scientific communication",
    "Time/project management", "Statistical analysis plans",
    "Professional correspondence", "Reproducibility",
    "Using strong statistical voice",
    "Databases, data sources, and data collection tools",
    "Learning new statistical methods", "Effective meeting strategy",
    "Collaboration with analytic colleagues",
    "Developing clinical/scientific domain knowledge",
    "Regulatory requirements", "Literature review",
    "Institutional structure"
  ),
  Early = c(89.9, 67.5, 74.7, 77.2, 68.4, 65.8, 58.2, 51.9, 57.0, 53.2, 39.2, 32.9, 27.8, 21.5, 15.2, 7.6),
  Mid = c(80.4, 80.4, 72.6, 72.6, 69.9, 67.3, 61.1, 61.1, 53.1, 51.3, 46.0, 38.9, 28.3, 23.0, 18.6, 11.5),
  Late = c(68.6, 83.3, 79.9, 68.1, 74.3, 66.4, 58.6, 60.0, 56.4, 45.7, 48.6, 42.1, 50.0, 25.7, 20.0, 18.6)
)

fig1_long <- fig1_data |>
  pivot_longer(cols = c(Early, Mid, Late), names_to = "Career_Stage", values_to = "Percent") |>
  mutate(
    Career_Stage = factor(Career_Stage, levels = c("Early", "Mid", "Late")),
    Skill = factor(Skill, levels = rev(fig1_data$Skill))
  )

ggplot(fig1_long, aes(x = Percent/100, y = Skill, fill = Career_Stage)) +
geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_fill_manual(values = c("Early" = "#619CFF", "Mid" = "#00BA38", "Late" = "#F8766D"),
                    name = "Career Stage") +
  labs(
    x = "Proportion Rating Skill As Absolutely Essential",
    y = NULL,
    title = "Skill Importance by Career Stage"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 11)
  )
```

---

## Figure 2: Skills by Highest Degree

```{r}
#| echo: false
#| message: false
#| fig-width: 14
#| fig-height: 7

fig2_data <- data.frame(
  Skill = c(
    "Coding", "Presenting results", "Scientific communication",
    "Time/project management", "Statistical analysis plans",
    "Professional correspondence", "Reproducibility",
    "Using strong statistical voice",
    "Databases, data sources, and data collection tools",
    "Learning new statistical methods", "Effective meeting strategy",
    "Collaboration with analytic colleagues",
    "Developing clinical/scientific domain knowledge",
    "Regulatory requirements", "Literature review",
    "Institutional structure"
  ),
  Masters = c(79.1, 79.1, 74.8, 74.1, 71.2, 66.2, 56.8, 54.0, 55.4, 41.9, 41.0, 33.1, 33.8, 25.9, 12.6, 6.7),
  Doctorate = c(76.5, 78.4, 76.0, 70.6, 73.0, 66.7, 60.8, 61.3, 55.4, 54.4, 48.0, 42.6, 39.7, 22.5, 22.4, 17.9)
)

fig2_long <- fig2_data |>
  pivot_longer(cols = c(Masters, Doctorate), names_to = "Degree", values_to = "Percent") |>
  mutate(
    Degree = factor(Degree, levels = c("Masters", "Doctorate")),
    Skill = factor(Skill, levels = rev(fig2_data$Skill))
  )

ggplot(fig2_long, aes(x = Percent/100, y = Skill, fill = Degree)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_fill_manual(values = c("Masters" = "#00BFC4", "Doctorate" = "#F8766D"),
                    name = "Highest Degree") +
  labs(
    x = "Proportion Rating Skill As Absolutely Essential",
    y = NULL,
    title = "Skill Importance by Highest Degree Earned"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 11)
  )
```

---

## Figure 3: Skills by Job Sector

```{r}
#| echo: false
#| message: false
#| fig-width: 14
#| fig-height: 7

fig3_data <- data.frame(
  Skill = c(
    "Scientific communication", "Presenting results", "Coding",
    "Statistical analysis plans", "Time/project management",
    "Professional correspondence", "Using strong statistical voice",
    "Databases, data sources, and data collection tools",
    "Reproducibility", "Effective meeting strategy",
    "Learning new statistical methods",
    "Developing clinical/scientific domain knowledge",
    "Collaboration with analytic colleagues",
    "Regulatory requirements", "Literature review",
    "Institutional structure"
  ),
  Academia = c(74.8, 80.6, 76.9, 71.8, 70.6, 66.0, 58.4, 51.3, 58.8, 44.5, 50.8, 38.7, 39.1, 19.6, 20.2, 14.7),
  Government = c(92.9, 80.0, 86.7, 93.3, 66.7, 66.7, 60.0, 60.0, 60.0, 46.7, 46.7, 40.0, 33.3, 26.7, 13.3, 6.7),
  Industry = c(74.3, 73.0, 79.7, 70.3, 74.3, 68.9, 59.5, 65.8, 58.1, 47.3, 44.6, 32.4, 37.8, 36.5, 13.5, 10.8),
  SelfEmployed = c(76.9, 84.6, 69.2, 69.2, 76.9, 61.5, 53.8, 53.8, 61.5, 46.2, 46.2, 38.5, 38.5, 23.1, 15.4, 7.7)
)

fig3_long <- fig3_data |>
  pivot_longer(cols = c(Academia, Government, Industry, SelfEmployed),
               names_to = "Sector", values_to = "Percent") |>
  mutate(
    Sector = factor(Sector, levels = c("Academia", "Government", "Industry", "SelfEmployed"),
                    labels = c("Academia", "Government", "Industry", "Self-employed")),
    Skill = factor(Skill, levels = rev(fig3_data$Skill))
  )

ggplot(fig3_long, aes(x = Percent/100, y = Skill, fill = Sector)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_fill_manual(values = c("Academia" = "#C77CFF", "Government" = "#619CFF",
                               "Industry" = "#00BA38", "Self-employed" = "#F8766D"),
                    name = "Job Sector") +
  labs(
    x = "Proportion Rating Skill As Absolutely Essential",
    y = NULL,
    title = "Skill Importance by Job Sector"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 11)
  )
```

---

## Key Findings

:::: {.columns}

::: {.column width="50%"}
**Top 5 Skills (% Absolutely Essential)**

1. Presenting results (78.6%)
2. Coding (77.6%)
3. Scientific communication (75.5%)
4. Statistical analysis plans (72.3%)
5. Time/project management (71.9%)
:::

::: {.column width="50%"}
**Key Differences**

- **Career stage**: Domain knowledge more valued by late-career; coding more valued by early-career
- **Industry vs Academia**: Regulatory requirements nearly 2× more essential in industry
- **Degree**: Learning new methods more essential for PhDs
:::

::::

---

## Reference

Slade E, Brearley AM, Coles A, Hayat MJ, Kulkarni PM, Nowacki AS, Oster RA, Posner MA, Samsa G, Spratt H, Troy J, and Pomann G-M.

**Essential team science skills for biostatisticians on collaborative research teams.**

*Journal of Clinical and Translational Science* 7: e243, 1–9.

doi: [10.1017/cts.2023.676](https://doi.org/10.1017/cts.2023.676)
