---
title: "Why NAMESPACE Cannot Replace renv"
subtitle: "Understanding Version Management in R"
author: "PHB 243b: Practicum in Biostatistics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# The Question

Can we use exact version specifications in `DESCRIPTION` instead of `renv`?

```
Imports:
    dplyr (== 1.1.4)
```

The short answer: **No.** This approach has fundamental limitations that make
it unsuitable as a replacement for `renv`.

# Where Version Information Lives

| File | Purpose | Example |
|------|---------|---------|
| `NAMESPACE` | Symbol binding | `importFrom(dplyr, filter)` |
| `DESCRIPTION` | Minimum version requirements | `Imports: dplyr (>= 1.1.0)` |
| `renv.lock` | Exact version pinning | `"dplyr": {"Version": "1.1.4"}` |

The `NAMESPACE` file cannot include version numbers at all---it only handles
symbol-level declarations (what functions to import/export).

# Why `==` in DESCRIPTION Falls Short

## Problem 1: Only Enforces Direct Dependencies

```
Imports:
    dplyr (== 1.1.4)
```

This pins `dplyr`, but what about dplyr's dependencies? The `dplyr` package
imports `vctrs`, `pillar`, `cli`, `rlang`, and others. Their versions remain
uncontrolled. A breaking change in `rlang` 1.2.0 vs 1.1.0 can still break
your analysis.

`renv.lock` captures the **entire dependency tree**---potentially dozens or
hundreds of packages with exact versions.

```{r dependency-tree, eval=FALSE}
# See the full dependency tree
pak::pkg_deps_tree("dplyr")
```

## Problem 2: No Installation Mechanism

`DESCRIPTION` only *declares* requirements; it does not *provide* old
versions. When you run `install.packages("dplyr")`, R fetches the current
CRAN version regardless of what `DESCRIPTION` says.

If dplyr 1.1.4 is no longer on CRAN (superseded by 1.2.0), installation
fails.

`renv` can:

- Restore from its local cache
- Install from the CRAN archive
- Install from a specified repository snapshot (e.g., Posit Public Package
  Manager)

```{r renv-restore, eval=FALSE}
# renv handles old versions automatically
renv::restore()
```

## Problem 3: Conflicts in Shared Environments

If two packages both specify exact versions:
```
Package A: dplyr (== 1.1.4)
Package B: dplyr (== 1.0.10)
```

Both cannot be satisfied simultaneously. Installation fails entirely.

`renv` creates **project-local libraries**, so each project can have its own
dplyr version without conflict.

```{r library-paths, eval=FALSE}
# renv isolates each project
.libPaths()
# [1] "/project/renv/library/R-4.4/aarch64-apple-darwin20"
# [2] "/Library/Frameworks/R.framework/.../library"
```

## Problem 4: R Version Not Captured

`DESCRIPTION` cannot specify the R version itself. An analysis run on R 4.3.2
may behave differently on R 4.5.0 (recall the `stringsAsFactors` change in
R 4.0.0).

`renv.lock` records the R version:
```json
{
  "R": {
    "Version": "4.4.2",
    "Repositories": [
      {
        "Name": "CRAN",
        "URL": "https://cloud.r-project.org"
      }
    ]
  }
}
```

Docker enforces the exact R version at runtime.

# Comparison Summary

| Capability | `DESCRIPTION` with `==` | `renv` |
|------------|:-----------------------:|:------:|
| Pin direct dependencies | Yes | Yes |
| Pin transitive dependencies | No | Yes |
| Install old versions | No | Yes |
| Project-local libraries | No | Yes |
| Record R version | No | Yes |
| Work across projects | Conflicts | Isolated |

# Conclusion

The `==` operator in `DESCRIPTION` is rarely used because it creates
brittleness without providing actual reproducibility. It declares a
requirement that the standard tooling cannot fulfill.

`renv` solves the problem that `==` only pretends to address:

- Full dependency tree capture
- Automated restoration of exact versions
- Project isolation
- Integration with repository snapshots and Docker

For reproducible research, use `DESCRIPTION` with `>=` for compatibility
bounds and `renv.lock` for exact reproducibility.
