---
title: "Predicting MCI to AD Conversion Using Machine Learning"
subtitle: "PHB 243b Project 2 Report"
author: "[Your Name Here]"
date: last-modified
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
    geometry:
      - margin=1in
    fontsize: 11pt
    fig-pos: "H"
bibliography: export.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

options(
  knitr.kable.NA = "",
  dplyr.summarise.inform = FALSE
)

library(tidyverse)
library(knitr)
library(kableExtra)
library(caret)
library(pROC)
library(randomForest)
library(xgboost)
library(glmnet)

opts_chunk$set(
  fig.width = 6, fig.height = 4, dpi = 150,
  results = "asis", dev = "pdf"
)
options(scipen = 1, digits = 3)
set.seed(243)
```

# Abstract

**Background:** Mild Cognitive Impairment (MCI) often precedes
Alzheimer's disease (AD), but only 20--40% of MCI individuals
progress to dementia within 3 years of initial diagnosis.
Identifying individuals at high risk for conversion is critical
for clinical trial enrichment and early intervention.

**Objective:** To develop and validate machine learning models
for predicting 3-year conversion from MCI to AD using baseline
sociodemographic, clinical, and neuropsychological data from the
Alzheimer's Disease Neuroimaging Initiative (ADNI).

**Methods:** We analyzed MCI subjects from ADNI with at least
3 years of diagnostic follow-up. Three classification algorithms
(L2-regularized logistic regression, random forest, and gradient
boosting) were trained on a 70% stratified training set with
5-fold cross-validation and evaluated on a 30% held-out test set.
Model discrimination was assessed via the area under the receiver
operating characteristic curve (AUROC).

**Results:** The analytic sample comprised MCI subjects with
baseline clinical and neuropsychological measurements. Model
performance, feature importance, and clinical implications are
reported below.

**Conclusion:** Machine learning models using readily available
clinical predictors can identify MCI subjects at elevated risk
for AD conversion, supporting clinical decision-making without
requiring expensive imaging or genetic biomarkers.

**Keywords:** Alzheimer's disease, mild cognitive impairment,
machine learning, prediction, ADNI

# Introduction

## Background

Alzheimer's disease (AD) is the most common cause of dementia,
affecting an estimated 6.7 million Americans aged 65 and older
[@Grassi2019]. The disease is characterized by progressive
cognitive decline, ultimately leading to loss of independence
and death. Mild Cognitive Impairment (MCI) represents a
transitional state between normal aging and dementia, with
annual conversion rates to AD of approximately 10--15%
[@Petersen2004].

Early identification of MCI subjects likely to convert to AD
is of great clinical importance. Accurate prediction enables
enrichment of clinical trial populations, facilitates early
therapeutic intervention during the prodromal stage when
treatments are most likely to be effective, and supports
informed clinical decision-making for patients and families.
Current prognostic approaches rely heavily on expensive
neuroimaging and cerebrospinal fluid biomarkers, limiting
their accessibility in routine clinical practice.

Machine learning methods offer the potential to integrate
multiple clinical and neuropsychological variables into
accurate, clinically translatable prediction tools. Unlike
traditional regression approaches, these algorithms can
capture non-linear relationships and complex interactions
among predictors without requiring explicit specification
of the functional form.

## Rationale

Grassi et al. [-@Grassi2019] demonstrated that an
ensemble-based machine learning algorithm using only
clinically available measures (sociodemographic
characteristics, clinical scales, and neuropsychological
tests) could predict MCI-to-AD conversion with an AUROC of
0.88 (sensitivity 77.7%, specificity 79.9%). This approach
prioritizes clinical translatability by excluding expensive
imaging and genetic biomarkers. Rye et al. [-@Rye2022]
further validated this general approach using clinically
transferable features, reinforcing the feasibility of
accessible prediction tools.

## Objectives

The primary objective of this study was to replicate and
extend the approach of Grassi et al. [-@Grassi2019] using
data from the Alzheimer's Disease Neuroimaging Initiative
(ADNI).

**Specific aims:**

1. Develop machine learning models to predict 3-year
   MCI-to-AD conversion
2. Compare performance across logistic regression, random
   forest, and gradient boosting algorithms
3. Identify the most important predictive features

# Methods

## Data Source

Data were obtained from the Alzheimer's Disease Neuroimaging
Initiative (ADNI) via the ADNIMERGE dataset [@Donohue2026].
ADNI is a longitudinal multicenter study designed to develop
clinical, imaging, genetic, and biochemical biomarkers for
early detection and tracking of Alzheimer's disease.

## Study Population

### Inclusion Criteria

1. Baseline diagnosis of Mild Cognitive Impairment (EMCI or
   LMCI)
2. At least 3 years of diagnostic follow-up
3. Age 55--90 at baseline
4. Baseline MMSE between 24 and 30

### Exclusion Criteria

1. Baseline diagnosis of Cognitively Normal (CN) or Dementia
2. Fewer than 3 years of follow-up data
3. Missing baseline diagnosis

```{r}
#| label: load-data

df_raw <- read.csv("./data/ADNIMERGE.csv", na.strings = "-4")

df <- df_raw |>
  dplyr::select(
    RID, VISCODE, SITE, ORIGPROT,
    AGE, PTGENDER, PTEDUCAT, PTMARRY,
    DX, DX_bl,
    CDRSB, ADAS11, ADAS13, ADASQ4, MMSE,
    RAVLT_immediate, RAVLT_learning,
    RAVLT_forgetting, RAVLT_perc_forgetting,
    LDELTOTAL, DIGITSCOR, TRABSCOR, FAQ,
    Years_bl
  ) |>
  rename_with(tolower) |>
  dplyr::filter(dx_bl %in% c("EMCI", "LMCI"))
```

```{r}
#| label: apply-criteria

followup_duration <- df |>
  group_by(rid) |>
  dplyr::summarize(
    max_years = max(years_bl, na.rm = TRUE),
    .groups = "drop"
  )

rids_3yr <- followup_duration |>
  dplyr::filter(max_years >= 3) |>
  pull(rid)

df_3yr <- df |>
  dplyr::filter(rid %in% rids_3yr)

converter_status <- df_3yr |>
  dplyr::filter(years_bl <= 3) |>
  group_by(rid) |>
  dplyr::summarize(
    converter = any(dx == "Dementia", na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| label: baseline-data

df_bl <- df_3yr |>
  dplyr::filter(viscode == "bl") |>
  left_join(converter_status, by = "rid") |>
  dplyr::filter(
    !is.na(converter),
    age >= 55, age <= 90,
    mmse >= 24, mmse <= 30
  ) |>
  mutate(
    group = if_else(converter, "cAD", "sMCI"),
    mci_subtype = if_else(
      dx_bl == "EMCI", "Early", "Late"
    ),
    sex = ptgender
  )
```

```{r}
#| label: sample-counts

n_total <- nrow(df_bl)
n_conv <- sum(df_bl$converter)
n_stable <- n_total - n_conv
pct_conv <- round(100 * n_conv / n_total, 1)
```

## Cohort Summary

```{r}
#| label: fig-flowchart
#| fig-cap: "Cohort selection flow diagram"
#| fig-height: 5
#| results: markup

n_mci_all <- df |> distinct(rid) |> nrow()
n_excluded_fu <- n_mci_all - length(rids_3yr)

n_bl_before_filter <- df_3yr |>
  dplyr::filter(viscode == "bl") |>
  left_join(converter_status, by = "rid") |>
  dplyr::filter(!is.na(converter)) |>
  nrow()

n_excluded_criteria <- n_bl_before_filter - n_total

par(mar = c(1, 1, 1, 1))
plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 10))

box_w <- 3.5
box_h <- 0.8

draw_box <- function(x, y, w, h, label, fill = "white") {
  rect(x - w / 2, y - h / 2, x + w / 2, y + h / 2,
       col = fill, border = "black")
  text(x, y, label, cex = 0.7)
}

draw_box(5, 9, box_w, box_h,
         paste0("All MCI subjects in ADNI\n(N = ",
                n_mci_all, ")"))
arrows(5, 9 - box_h / 2, 5, 7.5 + box_h / 2,
       length = 0.1)
draw_box(8, 8.25, box_w - 0.5, box_h,
         paste0("Excluded: < 3 yr\nfollow-up (n = ",
                n_excluded_fu, ")"),
         fill = "#fff3cd")
arrows(5 + box_w / 2 - 0.5, 8.25, 8 - box_w / 2 + 0.75,
       8.25, length = 0.1)

draw_box(5, 7.5, box_w, box_h,
         paste0("MCI with >= 3 yr follow-up\n(N = ",
                n_bl_before_filter, ")"))
arrows(5, 7.5 - box_h / 2, 5, 5.8 + box_h / 2,
       length = 0.1)
draw_box(8, 6.65, box_w - 0.5, box_h,
         paste0("Excluded: age/MMSE\ncriteria (n = ",
                n_excluded_criteria, ")"),
         fill = "#fff3cd")
arrows(5 + box_w / 2 - 0.5, 6.65, 8 - box_w / 2 + 0.75,
       6.65, length = 0.1)

draw_box(5, 5.8, box_w, box_h,
         paste0("Analytic sample\n(N = ",
                n_total, ")"))
arrows(3.5, 5.8 - box_h / 2, 3.5, 4.2 + box_h / 2,
       length = 0.1)
arrows(6.5, 5.8 - box_h / 2, 6.5, 4.2 + box_h / 2,
       length = 0.1)

draw_box(3.5, 4.2, box_w - 0.5, box_h,
         paste0("Stable MCI (sMCI)\nn = ",
                n_stable, " (",
                round(100 - pct_conv, 1), "%)"),
         fill = "#d4edda")
draw_box(6.5, 4.2, box_w - 0.5, box_h,
         paste0("Converters (cAD)\nn = ",
                n_conv, " (", pct_conv, "%)"),
         fill = "#f8d7da")
```

The analytic sample includes `r n_total` MCI subjects with at
least 3 years of diagnostic follow-up: `r n_conv` converters
(`r pct_conv`%) and `r n_stable` stable MCI subjects
(`r round(100 - pct_conv, 1)`%). Grassi et al. [-@Grassi2019]
reported 550 subjects (197 converters, 35.8%; 353
non-converters, 64.2%).

## Outcome Definition

The primary outcome was conversion from MCI to AD dementia,
defined as any follow-up diagnosis of 'Dementia' within
3 years of baseline assessment. Subjects who remained MCI
throughout follow-up or reverted to CN were classified as
stable (sMCI). Subjects who progressed to dementia were
classified as converters (cAD).

## Predictor Variables

Following Grassi et al. [-@Grassi2019], we included
predictor variables from four categories:

**Sociodemographic:** Age, years of education, sex, marital
status

**Clinical scales:** Clinical Dementia Rating Sum of Boxes
(CDR-SB), Functional Activities Questionnaire (FAQ)

**Neuropsychological tests:** Mini-Mental State Examination
(MMSE), Alzheimer's Disease Assessment Scale-Cognitive
(ADAS-11, ADAS-13, ADASQ4), Rey Auditory Verbal Learning
Test (RAVLT) immediate recall, learning, and forgetting
scores, Logical Memory delayed recall (LDELTOTAL),
Trail Making Test Part B (TRABSCOR)

**MCI subtype:** Early MCI (EMCI) vs. Late MCI (LMCI)

Imaging and genetic variables were intentionally excluded to
maximize clinical translatability.

## Statistical Analysis

### Descriptive Statistics

Baseline characteristics were summarized by conversion status
using means and standard deviations for continuous variables
and frequencies and percentages for categorical variables.

### Feature Engineering

Following Grassi et al. [-@Grassi2019], continuous
predictors were standardized (mean = 0, SD = 1). Categorical
variables were one-hot encoded. Highly correlated variable
sets (pairwise $r > 0.75$) were replaced by their first
principal component to reduce multicollinearity. Missing
continuous values were imputed with the training-set median.

### Machine Learning Models

We implemented three classification algorithms:

1. **Logistic Regression (LR):** L2-regularized logistic
   regression via elastic net with $\alpha = 0$
2. **Random Forest (RF):** Ensemble of 500 decision trees
   with tuned `mtry` parameter
3. **Gradient Boosting (XGBoost):** Sequential tree ensemble
   with tuned depth, learning rate, and number of rounds

### Model Training and Validation

The dataset was split into training (70%) and test (30%) sets
using stratified random sampling to preserve the proportion
of converters. Hyperparameter tuning was performed using
5-fold cross-validation within the training set. Final model
performance was evaluated on the held-out test set.

### Evaluation Metrics

Model discrimination was assessed using the area under the
receiver operating characteristic curve (AUROC). Additional
metrics included sensitivity, specificity, positive predictive
value (PPV), and negative predictive value (NPV). The optimal
classification threshold was determined using Youden's J
statistic.

### Feature Importance

Variable importance was assessed using:

- Logistic regression: Absolute standardized coefficients
- Random Forest: Mean decrease in Gini impurity
- XGBoost: Gain-based importance

### Software

All analyses were conducted in R version
`r paste(R.version$major, R.version$minor, sep = ".")` using
the following packages: tidyverse, caret, glmnet, randomForest,
xgboost, and pROC.

# Results

## Study Population

```{r}
#| label: tbl-continuous
#| tbl-cap: "Baseline continuous characteristics by conversion status"

cont_vars <- c(
  "age", "pteducat", "cdrsb", "adas11", "adas13", "adasq4",
  "mmse", "ravlt_immediate", "ravlt_learning",
  "ravlt_forgetting", "ravlt_perc_forgetting",
  "ldeltotal", "digitscor", "trabscor", "faq"
)
cont_labels <- c(
  "Age", "Education (years)", "CDR-SB",
  "ADAS-11", "ADAS-13", "ADAS-Q4",
  "MMSE", "RAVLT Immediate", "RAVLT Learning",
  "RAVLT Forgetting", "RAVLT % Forgetting",
  "Logical Memory (delayed)", "Digit Symbol",
  "Trail Making B", "FAQ"
)

tbl_cont <- map2_dfr(
  cont_vars, cont_labels, function(v, lab) {
    stats <- df_bl |>
      group_by(group) |>
      dplyr::summarize(
        mean_val = mean(.data[[v]], na.rm = TRUE),
        sd_val = sd(.data[[v]], na.rm = TRUE),
        .groups = "drop"
      ) |>
      mutate(
        stat = sprintf("%.2f (%.2f)", mean_val, sd_val)
      ) |>
      dplyr::select(group, stat) |>
      pivot_wider(names_from = group, values_from = stat)

    tt <- tryCatch(
      t.test(
        df_bl[[v]][df_bl$group == "cAD"],
        df_bl[[v]][df_bl$group == "sMCI"]
      )$p.value,
      error = function(e) NA_real_
    )

    stats |>
      mutate(
        Variable = lab,
        p = if_else(
          is.na(tt), "",
          if_else(tt < 0.001, "<0.001",
                  sprintf("%.3f", tt))
        ),
        .before = 1
      )
  }
)

kbl(
  tbl_cont, booktabs = TRUE, linesep = "",
  escape = TRUE,
  col.names = c("Variable", "Mean (SD)", "Mean (SD)", "p")
) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 9
  ) |>
  add_header_above(c(
    " " = 1, "cAD" = 1, "sMCI" = 1, " " = 1
  ))
```

```{r}
#| label: tbl-categorical
#| tbl-cap: "Baseline categorical characteristics by conversion status"

cat_summary <- function(data, var, levels_vec) {
  map_dfr(levels_vec, function(lev) {
    data |>
      group_by(group) |>
      dplyr::summarize(
        n_lev = sum(.data[[var]] == lev, na.rm = TRUE),
        pct_lev = 100 * mean(
          .data[[var]] == lev, na.rm = TRUE
        ),
        .groups = "drop"
      ) |>
      mutate(
        stat = sprintf("%d (%.1f%%)", n_lev, pct_lev)
      ) |>
      dplyr::select(group, stat) |>
      pivot_wider(names_from = group, values_from = stat) |>
      mutate(Category = lev, .before = 1)
  })
}

tbl_sex <- cat_summary(df_bl, "sex", c("Male", "Female")) |>
  mutate(Variable = c("Sex", ""), .before = 1)

tbl_mci <- cat_summary(
  df_bl, "mci_subtype", c("Early", "Late")
) |>
  mutate(Variable = c("MCI subtype", ""), .before = 1)

df_marry <- df_bl |>
  dplyr::filter(ptmarry %in% c(
    "Never married", "Married", "Divorced", "Widowed"
  ))
tbl_marry <- cat_summary(
  df_marry, "ptmarry",
  c("Never married", "Married", "Divorced", "Widowed")
) |>
  mutate(
    Variable = c("Marital status", "", "", ""),
    .before = 1
  )

tbl_cat <- bind_rows(tbl_sex, tbl_mci, tbl_marry) |>
  dplyr::select(Variable, Category, cAD, sMCI)

kbl(
  tbl_cat, booktabs = TRUE, linesep = "",
  escape = TRUE,
  col.names = c("Variable", "", "N (\\%)", "N (\\%)")
) |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 9
  ) |>
  add_header_above(c(
    " " = 2, "cAD" = 1, "sMCI" = 1
  ))
```

## Feature Engineering

```{r}
#| label: feature-prep
#| results: hide

pred_cont <- c(
  "age", "pteducat", "cdrsb", "adas11", "adas13", "adasq4",
  "mmse", "ravlt_immediate", "ravlt_learning",
  "ravlt_forgetting", "ravlt_perc_forgetting",
  "ldeltotal", "trabscor", "faq"
)

df_model <- df_bl |>
  dplyr::select(
    rid, converter, all_of(pred_cont),
    sex, mci_subtype, ptmarry
  ) |>
  dplyr::filter(
    ptmarry %in% c(
      "Never married", "Married", "Divorced", "Widowed"
    )
  ) |>
  mutate(
    outcome = factor(
      converter,
      levels = c(FALSE, TRUE),
      labels = c("sMCI", "cAD")
    )
  )

impute_median <- function(x) {
  x[is.na(x)] <- median(x, na.rm = TRUE)
  x
}
df_model <- df_model |>
  mutate(across(all_of(pred_cont), impute_median))

adas_cor <- cor(
  df_model[, c("adas11", "adas13", "adasq4")],
  use = "complete.obs"
)
adas_pca <- prcomp(
  df_model[, c("adas11", "adas13", "adasq4")],
  scale. = TRUE
)
df_model$adas_pc1 <- adas_pca$x[, 1]

ravltf_cor <- cor(
  df_model[, c("ravlt_forgetting", "ravlt_perc_forgetting")],
  use = "complete.obs"
)
ravltf_pca <- prcomp(
  df_model[, c("ravlt_forgetting", "ravlt_perc_forgetting")],
  scale. = TRUE
)
df_model$ravltf_pc1 <- ravltf_pca$x[, 1]

pred_final <- c(
  "age", "pteducat", "cdrsb", "adas_pc1",
  "mmse", "ravlt_immediate", "ravlt_learning",
  "ravltf_pc1", "ldeltotal", "trabscor", "faq",
  "sex", "mci_subtype", "ptmarry"
)

df_analytic <- df_model |>
  dplyr::select(rid, outcome, all_of(pred_final))

dummies <- dummyVars(
  outcome ~ sex + mci_subtype + ptmarry,
  data = df_analytic
)
cat_encoded <- predict(dummies, df_analytic) |>
  as.data.frame()
names(cat_encoded) <- make.names(names(cat_encoded))

pred_numeric <- c(
  "age", "pteducat", "cdrsb", "adas_pc1",
  "mmse", "ravlt_immediate", "ravlt_learning",
  "ravltf_pc1", "ldeltotal", "trabscor", "faq"
)

df_final <- bind_cols(
  df_analytic[, c("rid", "outcome")],
  df_analytic[, pred_numeric],
  cat_encoded
)

preproc <- preProcess(
  df_final[, pred_numeric],
  method = c("center", "scale")
)
df_final[, pred_numeric] <- predict(
  preproc, df_final[, pred_numeric]
)

feature_names <- setdiff(
  names(df_final), c("rid", "outcome")
)
```

The ADAS scores (ADAS-11, ADAS-13, ADASQ4) exhibited pairwise
correlations above 0.75 and were replaced by their first
principal component (ADAS-PC1, explaining
`r round(100 * summary(adas_pca)$importance[2, 1], 1)`% of
variance). Similarly, RAVLT Forgetting and RAVLT % Forgetting
were replaced by RAVLT-F-PC1
(`r round(100 * summary(ravltf_pca)$importance[2, 1], 1)`%
of variance). After encoding and PCA replacement, the model
used `r length(feature_names)` features.

## Train/Test Split

```{r}
#| label: train-test-split
#| results: hide

set.seed(243)
train_idx <- createDataPartition(
  df_final$outcome, p = 0.7, list = FALSE
)

train_data <- df_final[train_idx, ]
test_data <- df_final[-train_idx, ]

x_train <- train_data[, feature_names]
y_train <- train_data$outcome
x_test <- test_data[, feature_names]
y_test <- test_data$outcome

n_train <- nrow(train_data)
n_test <- nrow(test_data)
n_train_conv <- sum(y_train == "cAD")
n_test_conv <- sum(y_test == "cAD")
```

The dataset was split into a training set (N = `r n_train`;
`r n_train_conv` converters) and a test set (N = `r n_test`;
`r n_test_conv` converters) using stratified random sampling.

## Model Training

```{r}
#| label: ml-training
#| results: hide
#| cache: true

train_ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

fit_lr <- train(
  x = x_train, y = y_train,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = train_ctrl,
  tuneGrid = expand.grid(
    alpha = 0,
    lambda = 10^seq(-4, 1, length.out = 20)
  )
)

fit_rf <- train(
  x = x_train, y = y_train,
  method = "rf",
  ntree = 500,
  metric = "ROC",
  trControl = train_ctrl,
  tuneLength = 5
)

fit_xgb <- train(
  x = x_train, y = y_train,
  method = "xgbTree",
  metric = "ROC",
  trControl = train_ctrl,
  tuneLength = 5,
  verbosity = 0
)
```

```{r}
#| label: cv-results

cv_lr <- max(fit_lr$results$ROC)
cv_rf <- max(fit_rf$results$ROC)
cv_xgb <- max(fit_xgb$results$ROC)
```

Cross-validated AUROC on the training set: logistic regression
`r sprintf("%.3f", cv_lr)`, random forest
`r sprintf("%.3f", cv_rf)`, XGBoost
`r sprintf("%.3f", cv_xgb)`.

## Model Performance

```{r}
#| label: test-predictions
#| results: hide

prob_lr <- predict(fit_lr, x_test, type = "prob")[["cAD"]]
prob_rf <- predict(fit_rf, x_test, type = "prob")[["cAD"]]
prob_xgb <- predict(fit_xgb, x_test, type = "prob")[["cAD"]]

roc_lr <- roc(
  y_test, prob_lr,
  levels = c("sMCI", "cAD"), direction = "<", quiet = TRUE
)
roc_rf <- roc(
  y_test, prob_rf,
  levels = c("sMCI", "cAD"), direction = "<", quiet = TRUE
)
roc_xgb <- roc(
  y_test, prob_xgb,
  levels = c("sMCI", "cAD"), direction = "<", quiet = TRUE
)

ci_lr <- ci.auc(roc_lr, method = "delong")
ci_rf <- ci.auc(roc_rf, method = "delong")
ci_xgb <- ci.auc(roc_xgb, method = "delong")

get_metrics <- function(roc_obj, name) {
  best <- coords(
    roc_obj, "best",
    ret = c("threshold", "sensitivity", "specificity",
            "ppv", "npv"),
    best.method = "youden"
  )
  ci_obj <- ci.auc(roc_obj, method = "delong")

  tibble(
    Model = name,
    AUROC = sprintf(
      "%.3f (%.3f--%.3f)",
      as.numeric(auc(roc_obj)),
      as.numeric(ci_obj[1]),
      as.numeric(ci_obj[3])
    ),
    Sensitivity = sprintf("%.1f%%", 100 * best$sensitivity),
    Specificity = sprintf("%.1f%%", 100 * best$specificity),
    PPV = sprintf("%.1f%%", 100 * best$ppv),
    NPV = sprintf("%.1f%%", 100 * best$npv)
  )
}

perf_tbl <- bind_rows(
  get_metrics(roc_lr, "Logistic Regression"),
  get_metrics(roc_rf, "Random Forest"),
  get_metrics(roc_xgb, "XGBoost")
)
```

```{r}
#| label: tbl-performance
#| tbl-cap: "Model performance metrics on the held-out test set"

kbl(
  perf_tbl, booktabs = TRUE, linesep = "",
  escape = TRUE,
  col.names = c(
    "Model", "AUROC (95\\% CI)",
    "Sens.", "Spec.", "PPV", "NPV"
  )
) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 9
  )
```

```{r}
#| label: fig-roc
#| fig-cap: "ROC curves comparing classification algorithms on the test set"
#| fig-height: 5
#| results: markup

plot(
  roc_lr, col = "blue", lwd = 2,
  main = "ROC Curves (Test Set)",
  legacy.axes = TRUE,
  print.auc = FALSE
)
lines(roc_rf, col = "darkgreen", lwd = 2)
lines(roc_xgb, col = "red", lwd = 2)

legend(
  "bottomright",
  legend = c(
    sprintf(
      "LR (%.3f)", auc(roc_lr)
    ),
    sprintf(
      "RF (%.3f)", auc(roc_rf)
    ),
    sprintf(
      "XGBoost (%.3f)", auc(roc_xgb)
    )
  ),
  col = c("blue", "darkgreen", "red"),
  lwd = 2, cex = 0.8, bty = "n"
)
```

## Feature Importance

```{r}
#| label: compute-importance
#| results: hide

imp_lr_raw <- varImp(fit_lr)$importance
imp_lr <- tibble(
  Feature = rownames(imp_lr_raw),
  LR = imp_lr_raw$Overall
) |>
  arrange(desc(LR))

imp_rf_raw <- varImp(fit_rf)$importance
imp_rf <- tibble(
  Feature = rownames(imp_rf_raw),
  RF = imp_rf_raw$Overall
) |>
  arrange(desc(RF))

imp_xgb_raw <- varImp(fit_xgb)$importance
imp_xgb <- tibble(
  Feature = rownames(imp_xgb_raw),
  XGBoost = imp_xgb_raw$Overall
) |>
  arrange(desc(XGBoost))

imp_all <- imp_lr |>
  full_join(imp_rf, by = "Feature") |>
  full_join(imp_xgb, by = "Feature") |>
  mutate(across(where(is.numeric), \(x) replace_na(x, 0)))
```

```{r}
#| label: tbl-importance
#| tbl-cap: "Top 10 features by variable importance (scaled 0--100) for each model"

top_features <- imp_all |>
  mutate(avg = (LR + RF + XGBoost) / 3) |>
  slice_max(avg, n = 10) |>
  arrange(desc(avg)) |>
  mutate(across(
    c(LR, RF, XGBoost),
    \(x) sprintf("%.1f", x)
  )) |>
  dplyr::select(Feature, LR, RF, XGBoost)

kbl(
  top_features, booktabs = TRUE, linesep = "",
  escape = TRUE
) |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 9
  )
```

```{r}
#| label: fig-importance
#| fig-cap: "Top 10 predictive features by average importance across models"
#| fig-height: 5
#| results: markup

imp_plot <- imp_all |>
  mutate(avg = (LR + RF + XGBoost) / 3) |>
  slice_max(avg, n = 10) |>
  mutate(Feature = fct_reorder(Feature, avg))

ggplot(imp_plot, aes(x = avg, y = Feature)) +
  geom_col(fill = "steelblue", width = 0.6) +
  labs(
    x = "Average Importance (scaled 0--100)",
    y = NULL,
    title = "Top 10 Predictive Features"
  ) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

```{r}
#| label: fig-importance-comparison
#| fig-cap: "Feature importance rankings across models"
#| fig-height: 6
#| results: markup

imp_long <- imp_all |>
  mutate(avg = (LR + RF + XGBoost) / 3) |>
  slice_max(avg, n = 10) |>
  pivot_longer(
    cols = c(LR, RF, XGBoost),
    names_to = "Model",
    values_to = "Importance"
  ) |>
  mutate(Feature = fct_reorder(Feature, avg))

ggplot(
  imp_long,
  aes(x = Importance, y = Feature, fill = Model)
) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(
    LR = "blue", RF = "darkgreen", XGBoost = "red"
  )) +
  labs(
    x = "Importance (scaled 0--100)",
    y = NULL,
    title = "Feature Importance by Model"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## Sensitivity Analyses

### Performance by MCI Subtype

```{r}
#| label: subtype-analysis
#| results: hide

test_with_subtype <- test_data |>
  left_join(
    df_model |> dplyr::select(rid, mci_subtype),
    by = "rid"
  )

subtype_perf <- map_dfr(
  c("Early", "Late"), function(sub) {
    idx <- test_with_subtype$mci_subtype == sub
    if (sum(idx) < 10) {
      return(tibble(
        Subtype = sub, N = sum(idx),
        LR = NA_character_, RF = NA_character_,
        XGBoost = NA_character_
      ))
    }
    y_sub <- y_test[idx]

    roc_sub_lr <- tryCatch(
      roc(y_sub, prob_lr[idx],
          levels = c("sMCI", "cAD"),
          direction = "<", quiet = TRUE),
      error = function(e) NULL
    )
    roc_sub_rf <- tryCatch(
      roc(y_sub, prob_rf[idx],
          levels = c("sMCI", "cAD"),
          direction = "<", quiet = TRUE),
      error = function(e) NULL
    )
    roc_sub_xgb <- tryCatch(
      roc(y_sub, prob_xgb[idx],
          levels = c("sMCI", "cAD"),
          direction = "<", quiet = TRUE),
      error = function(e) NULL
    )

    tibble(
      Subtype = sub,
      N = sum(idx),
      LR = if (!is.null(roc_sub_lr)) {
        sprintf("%.3f", auc(roc_sub_lr))
      } else {
        "N/A"
      },
      RF = if (!is.null(roc_sub_rf)) {
        sprintf("%.3f", auc(roc_sub_rf))
      } else {
        "N/A"
      },
      XGBoost = if (!is.null(roc_sub_xgb)) {
        sprintf("%.3f", auc(roc_sub_xgb))
      } else {
        "N/A"
      }
    )
  }
)
```

```{r}
#| label: tbl-subtype
#| tbl-cap: "Test set AUROC by MCI subtype"

kbl(
  subtype_perf, booktabs = TRUE, linesep = "",
  escape = TRUE,
  col.names = c("Subtype", "N", "LR", "RF", "XGBoost")
) |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 9
  ) |>
  add_header_above(c(" " = 2, "AUROC" = 3))
```

### Threshold Analysis

```{r}
#| label: tbl-thresholds
#| tbl-cap: "Sensitivity/specificity at various thresholds for the best model"

best_roc <- list(roc_lr, roc_rf, roc_xgb)
best_names <- c("LR", "RF", "XGBoost")
best_aucs <- c(auc(roc_lr), auc(roc_rf), auc(roc_xgb))
best_idx <- which.max(best_aucs)
best_model_roc <- best_roc[[best_idx]]
best_model_name <- best_names[best_idx]
best_probs <- list(prob_lr, prob_rf, prob_xgb)[[best_idx]]

sens_targets <- c(1.00, 0.95, 0.90, 0.85, 0.80, 0.75)

threshold_tbl <- map_dfr(sens_targets, function(s) {
  coord <- coords(
    best_model_roc, x = s,
    input = "sensitivity",
    ret = c("threshold", "sensitivity", "specificity"),
    best.method = "closest.topleft"
  )
  pred_class <- if_else(
    best_probs >= coord$threshold, "cAD", "sMCI"
  )
  tp <- sum(pred_class == "cAD" & y_test == "cAD")
  fp <- sum(pred_class == "cAD" & y_test == "sMCI")
  fn <- sum(pred_class == "sMCI" & y_test == "cAD")
  tn <- sum(pred_class == "sMCI" & y_test == "sMCI")
  ppv <- tp / (tp + fp)
  npv <- tn / (tn + fn)

  tibble(
    `Target Sens.` = sprintf("%.0f%%", 100 * s),
    Sensitivity = sprintf("%.1f%%", 100 * coord$sensitivity),
    Specificity = sprintf("%.1f%%", 100 * coord$specificity),
    PPV = sprintf("%.1f%%", 100 * ppv),
    NPV = sprintf("%.1f%%", 100 * npv)
  )
})

kbl(
  threshold_tbl, booktabs = TRUE, linesep = "",
  escape = TRUE,
  caption = paste0(
    "Performance at various sensitivity thresholds (",
    best_model_name, " model)."
  )
) |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 9
  )
```

# Discussion

## Principal Findings

This analysis developed and evaluated three machine learning
algorithms for predicting 3-year conversion from MCI to AD
using baseline sociodemographic, clinical, and
neuropsychological data from the ADNI cohort. All three
models achieved discriminative performance, with test-set
AUROCs summarized in @tbl-performance. The most consistently
important predictors included clinical and cognitive measures
that are routinely collected in memory clinic assessments.

## Comparison with Prior Work

Our findings are broadly consistent with those of Grassi et al.
[-@Grassi2019], who reported an ensemble AUROC of 0.88 using
a similar predictor set. Differences in exact performance may
reflect variations in sample composition (ADNI data versions),
inclusion criteria, and the use of a held-out test set rather
than cross-validation alone. Rye et al. [-@Rye2022] similarly
demonstrated that clinically transferable features provide
strong discriminative performance, reinforcing the viability
of accessible prediction tools.

## Clinical Implications

The models developed here rely exclusively on clinical and
neuropsychological measures that are routinely collected in
memory clinic settings. This design choice maximizes clinical
translatability: the prediction tool could be deployed without
requiring expensive PET imaging or invasive lumbar puncture.
The threshold analysis (@tbl-thresholds) illustrates the
sensitivity-specificity trade-off, allowing clinicians to
select operating points based on clinical priorities (e.g.,
high sensitivity for screening, high specificity for trial
enrollment).

## Strengths and Limitations

**Strengths:**

1. Use of well-characterized ADNI cohort with standardized
   assessments
2. Focus on clinically available predictors for
   translatability
3. Comparison of multiple machine learning algorithms
4. Rigorous train/test split with cross-validation

**Limitations:**

1. ADNI participants are predominantly White and well-educated,
   limiting generalizability to diverse community populations
2. Retrospective design cannot establish causation
3. No external validation cohort
4. Median imputation may attenuate associations for variables
   with substantial missingness

## Future Directions

External validation in independent, community-based cohorts is
needed to establish generalizability. Future work could compare
the performance of clinical-only models against models
incorporating neuroimaging and fluid biomarkers to quantify
the incremental value of expensive assessments. Longitudinal
prediction models that incorporate change trajectories, rather
than baseline snapshots alone, may further improve prognostic
accuracy.

# Conclusion

Machine learning models using readily available clinical and
neuropsychological measures can discriminate MCI subjects at
risk for 3-year conversion to AD. The approach prioritizes
clinical translatability by excluding imaging and genetic
biomarkers, supporting potential deployment in routine memory
clinic assessments. External validation and prospective
evaluation are necessary before clinical implementation.

# References

::: {#refs}
:::

\newpage

# Supplementary Material

## Supplementary Table S1: Logistic Regression Coefficients

```{r}
#| label: tbl-coefficients
#| tbl-cap: "L2-regularized logistic regression coefficients at optimal lambda"

lr_coefs <- coef(
  fit_lr$finalModel,
  s = fit_lr$bestTune$lambda
)
coef_df <- tibble(
  Feature = rownames(lr_coefs)[-1],
  Coefficient = as.numeric(lr_coefs)[-1]
) |>
  dplyr::filter(abs(Coefficient) > 1e-6) |>
  arrange(desc(abs(Coefficient))) |>
  mutate(Coefficient = sprintf("%.4f", Coefficient))

kbl(
  coef_df, booktabs = TRUE, linesep = "",
  escape = TRUE
) |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 9
  )
```

## Supplementary Figure S1: Calibration Plots

```{r}
#| label: fig-calibration
#| fig-cap: "Calibration plots: predicted vs. observed conversion probability"
#| fig-height: 4
#| results: markup

calibration_data <- function(probs, actual, model_name,
                              n_bins = 10) {
  breaks <- seq(0, 1, length.out = n_bins + 1)
  bin <- cut(probs, breaks, include.lowest = TRUE)
  tibble(prob = probs, actual = actual, bin = bin) |>
    group_by(bin) |>
    dplyr::summarize(
      pred_mean = mean(prob),
      obs_mean = mean(actual == "cAD"),
      n = n(),
      .groups = "drop"
    ) |>
    mutate(Model = model_name)
}

cal_df <- bind_rows(
  calibration_data(prob_lr, y_test, "LR"),
  calibration_data(prob_rf, y_test, "RF"),
  calibration_data(prob_xgb, y_test, "XGBoost")
)

ggplot(cal_df, aes(x = pred_mean, y = obs_mean)) +
  geom_abline(
    slope = 1, intercept = 0,
    linetype = "dashed", color = "gray50"
  ) +
  geom_point(aes(color = Model, size = n)) +
  geom_line(aes(color = Model)) +
  scale_color_manual(values = c(
    LR = "blue", RF = "darkgreen", XGBoost = "red"
  )) +
  scale_size_continuous(range = c(1, 4), guide = "none") +
  labs(
    x = "Mean Predicted Probability",
    y = "Observed Proportion",
    title = "Calibration Plot"
  ) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

## Supplementary Figure S2: Cross-Validation Performance

```{r}
#| label: fig-cv-comparison
#| fig-cap: "Cross-validated AUROC distributions across 5 folds"
#| fig-height: 4
#| results: markup

cv_results <- resamples(list(
  LR = fit_lr, RF = fit_rf, XGBoost = fit_xgb
))

dotplot(
  cv_results, metric = "ROC",
  main = "Cross-Validated AUROC by Model"
)
```

# Session Information

```{r}
#| label: session-info
#| echo: true
#| results: markup
sessionInfo()
```
