---
title: "Predicting MCI to AD Conversion Using Machine Learning"
subtitle: "PHB 243b Project 2 Report"
author: "[Your Name Here]"
date: last-modified
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
    geometry:
      - margin=1in
    fontsize: 11pt
    fig-pos: "H"
bibliography: export.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(knitr)
library(pROC)
library(caret)
# library(ADNIMERGE2)  # Uncomment when data available

# Set global options
opts_chunk$set(fig.width = 6, fig.height = 4, dpi = 150)
```

# Abstract

<!--
PROMPT: Write a structured abstract (250 words max) with:
- Background (1-2 sentences)
- Objective (1 sentence)
- Methods (2-3 sentences)
- Results (2-3 sentences with key numbers)
- Conclusion (1-2 sentences)
-->

**Background:** Mild Cognitive Impairment (MCI) often precedes Alzheimer's
disease (AD), but not all MCI patients progress to dementia. Identifying
individuals at high risk for conversion is critical for clinical management
and trial enrollment.

**Objective:** To develop and validate machine learning models for predicting
3-year conversion from MCI to AD using baseline clinical and cognitive data.

**Methods:** [Describe your methods in 2-3 sentences]

**Results:** [Report key findings: sample size, best model AUROC, top
predictors]

**Conclusion:** [Summarize clinical implications in 1-2 sentences]

**Keywords:** Alzheimer's disease, mild cognitive impairment, machine learning,
prediction, ADNI

# Introduction

## Background

Alzheimer's disease (AD) is the most common cause of dementia, affecting
an estimated 6.7 million Americans aged 65 and older [@Grassi2019]. The
disease is characterized by progressive cognitive decline, ultimately
leading to loss of independence and death. Mild Cognitive Impairment (MCI)
represents a transitional state between normal aging and dementia, with
annual conversion rates to AD of approximately 10-15% [@Petersen2004].

<!--
PROMPT: Expand on:
- The clinical importance of early prediction
- Current approaches to MCI diagnosis and prognosis
- Limitations of existing prediction methods
- The potential of machine learning approaches
-->

[Your expanded background here - 2-3 paragraphs]

## Rationale

Grassi et al. (2019) demonstrated that machine learning algorithms using
only clinically-available measures (sociodemographic characteristics,
clinical scales, and neuropsychological tests) could predict MCI-to-AD
conversion with an AUROC of 0.88 [@Grassi2019]. This approach prioritizes
clinical translatability over maximum predictive accuracy, excluding
expensive imaging and genetic biomarkers.

## Objectives

The primary objective of this study was to replicate and extend the
approach of Grassi et al. (2019) using data from the Alzheimer's Disease
Neuroimaging Initiative (ADNI).

**Specific aims:**

1. Develop machine learning models to predict 3-year MCI-to-AD conversion
2. Compare performance across multiple algorithms
3. Identify the most important predictive features

# Methods

## Data Source

Data were obtained from the Alzheimer's Disease Neuroimaging Initiative
(ADNI) via the ADNIMERGE2 R package [@Donohue2026]. ADNI is a longitudinal
multicenter study designed to develop clinical, imaging, genetic, and
biochemical biomarkers for early detection and tracking of Alzheimer's
disease. Data were downloaded on [DATE].

## Study Population

### Inclusion Criteria

1. Baseline diagnosis of Mild Cognitive Impairment (EMCI or LMCI)
2. At least one follow-up visit with diagnostic assessment
3. Complete data on primary predictor variables

### Exclusion Criteria

1. Baseline diagnosis of Cognitively Normal (CN) or Dementia
2. No follow-up visits after baseline

```{r}
#| label: fig-flowchart
#| fig-cap: "Cohort selection flow diagram"
#| eval: false

# Create flow diagram showing:
# - Total MCI subjects
# - Exclusions (with reasons)
# - Final analytic sample
# - Converters vs. Stable
```

## Outcome Definition

The primary outcome was conversion from MCI to AD dementia, defined as
any follow-up diagnosis of "Dementia" within 3 years of baseline assessment.
Subjects who remained MCI throughout follow-up or reverted to CN were
classified as stable (sMCI). Subjects who progressed to dementia were
classified as converters (cAD).

## Predictor Variables

Following Grassi et al. (2019), we included predictor variables from
four categories:

**Sociodemographic:** Age, years of education, sex, marital status

**Clinical scales:** Clinical Dementia Rating Sum of Boxes (CDR-SB),
Functional Activities Questionnaire (FAQ)

**Neuropsychological tests:** Mini-Mental State Examination (MMSE),
Alzheimer's Disease Assessment Scale-Cognitive 13-item (ADAS-Cog-13),
Rey Auditory Verbal Learning Test (RAVLT) immediate recall, learning,
and forgetting scores, Logical Memory delayed recall (LDELTOTAL),
Trail Making Test Part B (TMT-B)

**MCI subtype:** Early MCI (EMCI) vs. Late MCI (LMCI)

Imaging and genetic variables were intentionally excluded to maximize
clinical translatability.

## Statistical Analysis

### Descriptive Statistics

Baseline characteristics were summarized by conversion status using means
and standard deviations for continuous variables and frequencies and
percentages for categorical variables. Group comparisons used t-tests
for continuous variables and chi-square tests for categorical variables.

### Machine Learning Models

We implemented three classification algorithms:

1. **Logistic Regression:** Baseline linear classifier with L2 regularization
2. **Random Forest:** Ensemble of decision trees with 500 trees
3. **Gradient Boosting (XGBoost):** Sequential tree ensemble

### Model Training and Validation

The dataset was split into training (70%) and test (30%) sets using
stratified random sampling to preserve the proportion of converters.
Hyperparameter tuning was performed using 5-fold cross-validation within
the training set. Final model performance was evaluated on the held-out
test set.

### Evaluation Metrics

Model discrimination was assessed using the area under the receiver
operating characteristic curve (AUROC). Additional metrics included
sensitivity, specificity, positive predictive value (PPV), and negative
predictive value (NPV). The optimal classification threshold was determined
using Youden's J statistic.

### Feature Importance

Variable importance was assessed using:

- Logistic regression: Standardized coefficients
- Random Forest: Mean decrease in Gini impurity
- XGBoost: Gain-based importance

### Software

All analyses were conducted in R version 4.4.x using the following packages:
tidyverse, caret, randomForest, xgboost, pROC, and ADNIMERGE2.

# Results

## Study Population

```{r}
#| label: tbl-characteristics
#| tbl-cap: "Baseline characteristics by conversion status"
#| eval: false

# Create Table 1 showing:
# - Demographics (age, education, sex)
# - Clinical measures (MMSE, ADAS-Cog-13, CDR-SB, FAQ)
# - Neuropsychological tests (RAVLT, LDELTOTAL, TMT-B)
# - MCI subtype (EMCI vs LMCI)
# Stratified by sMCI vs cAD with p-values
```

<!--
PROMPT: Describe the study sample:
- Total N and conversion rate
- Key demographic differences between groups
- Key clinical/cognitive differences
-->

[Your results description here]

## Model Performance

```{r}
#| label: fig-roc
#| fig-cap: "ROC curves comparing classification algorithms"
#| eval: false

# Create overlaid ROC curves for all models
# Include AUROC values in legend
```

```{r}
#| label: tbl-performance
#| tbl-cap: "Model performance metrics on test set"
#| eval: false

# Create table showing:
# - AUROC (95% CI)
# - Sensitivity
# - Specificity
# - PPV
# - NPV
# - Accuracy
# For each model
```

<!--
PROMPT: Describe model performance:
- Which model performed best?
- How does performance compare to Grassi et al. (2019)?
- Sensitivity/specificity at optimal threshold
-->

[Your model performance results here]

## Feature Importance

```{r}
#| label: fig-importance
#| fig-cap: "Top 10 predictive features by importance"
#| eval: false

# Create bar chart or dot plot showing feature importance
# For best-performing model
```

```{r}
#| label: tbl-importance
#| tbl-cap: "Feature importance rankings across models"
#| eval: false

# Create table showing top 5-10 features for each model
```

<!--
PROMPT: Describe feature importance:
- Which variables were most predictive?
- Was importance consistent across models?
- How does this compare to Grassi et al. findings?
-->

[Your feature importance results here]

## Sensitivity Analyses

<!--
PROMPT: Report results of sensitivity analyses:
- EMCI-only analysis
- LMCI-only analysis
- Alternative model specifications
-->

[Your sensitivity analysis results here]

# Discussion

## Principal Findings

<!--
PROMPT: Summarize key findings (1-2 paragraphs):
- Best model and its performance
- Most important predictors
- Comparison to prior work
-->

[Your principal findings here]

## Comparison with Prior Work

Our findings are [consistent with / different from] those of Grassi et al.
(2019), who reported an AUROC of 0.88 using a similar approach. [Discuss
similarities and differences in performance, feature importance, etc.]

<!--
PROMPT: Compare your results to:
- Grassi et al. (2019)
- Rye et al. (2022)
- Other relevant studies
-->

[Your comparison discussion here]

## Clinical Implications

<!--
PROMPT: Discuss practical applications:
- How could this model be used in clinical practice?
- What are the implications of the sensitivity/specificity trade-off?
- Who would benefit most from this prediction tool?
-->

[Your clinical implications discussion here]

## Strengths and Limitations

**Strengths:**

1. Use of well-characterized ADNI cohort with standardized assessments
2. Focus on clinically-available predictors for translatability
3. Comparison of multiple machine learning algorithms
4. Rigorous cross-validation methodology

**Limitations:**

1. ADNI population may not generalize to community settings
2. Retrospective design cannot establish causation
3. No external validation cohort
4. [Additional limitations specific to your analysis]

## Future Directions

<!--
PROMPT: Suggest next steps:
- External validation in other cohorts
- Addition of biomarkers for comparison
- Longitudinal prediction models
- Implementation studies
-->

[Your future directions here]

# Conclusion

<!--
PROMPT: Write a concise conclusion (1 paragraph):
- Restate main finding
- Emphasize clinical relevance
- Note key limitation
- Suggest implication for practice or research
-->

[Your conclusion here]

# References

::: {#refs}
:::

# Supplementary Material

## Supplementary Table S1: Complete Model Coefficients

```{r}
#| label: tbl-coefficients
#| tbl-cap: "Logistic regression coefficients with 95% CI"
#| eval: false

# Full coefficient table from logistic regression
```

## Supplementary Figure S1: Calibration Plots

```{r}
#| label: fig-calibration
#| fig-cap: "Calibration plots for each model"
#| eval: false

# Predicted vs observed probability plots
```

## Supplementary Figure S2: Learning Curves

```{r}
#| label: fig-learning
#| fig-cap: "Learning curves showing training vs validation performance"
#| eval: false

# Plot showing performance vs training set size
```

# Session Information

```{r}
#| label: session-info
#| echo: true
sessionInfo()
```

<!--
CHECKLIST before submitting:

[ ] Abstract complete with all required sections
[ ] Introduction provides adequate background and rationale
[ ] Methods fully reproducible (another analyst could replicate)
[ ] Results presented clearly with appropriate tables/figures
[ ] Discussion interprets findings in context of prior work
[ ] Limitations honestly acknowledged
[ ] All figures have captions and are referenced in text
[ ] All tables have titles and are referenced in text
[ ] References properly formatted
[ ] Code runs without errors
[ ] Document renders to PDF without errors
-->
